# Module 4: PrognosiX Extended Workflow

This extended `PrognosiX` workflow reflects the package's newer prognosis capabilities and supports full clinical prognostic-model development and validation. It adds competing risks, time-dependent ROC comparison, multi-time calibration, repeated cross-validation, C-index comparison, Bayesian tuning, AIC-guided selection, overall calibration, external-validation diagnostics, and deployment.

## Execution Notes

This workflow is computationally intensive and uses optional packages that are checked only when their methods are requested. Verify paths, outcome definitions, prediction horizons, the output root, and dependencies before running. Keep development and validation independent throughout, report uncertainty with point estimates, and distinguish exploratory comparisons from prespecified confirmatory analyses.

## Step-by-step Workflow

<!-- split-into-step-chunks -->

The workflow is divided into short stages. Each note explains the purpose, key checks, and interpretation limits. Generated plots and tables appear directly below their code chunk.

### Environment and Global Setup

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module4-pro-01
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# =============================================================================
# icare Package — Module 4: Survival / Prognosis Modeling (PrognosiX)
# ADVANCED / PUBLICATION EDITION — STRICT TRAIN/VALIDATION SPLIT
# =============================================================================
#
# This is a complete, production‑ready pipeline for developing and validating
# a clinical survival model. It includes 10 advanced extensions covering
# model comparison, competing risks, multi‑time calibration, Bayesian tuning,
# clinical impact, and more – suitable for high‑impact journals.
#
# Pipeline structure:
````

### 0 Environment setup and path management

This section defines seeds, output paths, and dependencies. Update these paths and key parameters first when using new data.

````{r}
#| label: module4-pro-02
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   0. Environment setup and path management
````

### 1 Data preparation (split time/status into info.data)

This section imports and organizes the data. Check dimensions and variable coding before continuing.

````{r}
#| label: module4-pro-03
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   1. Data preparation (split time/status into info.data)
````

### 2 Feature selection (multi‑method, training set only)

This section selects candidate features. Consider effect size, stability, and biological plausibility rather than a single significance measure.

````{r}
#| label: module4-pro-04
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   2. Feature selection (multi‑method, training set only)
````

### 3 Algorithm benchmarking and hyperparameter tuning

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-pro-05
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   3. Algorithm benchmarking and hyperparameter tuning
````

### 4 Core model evaluation (KM, forest, time‑dependent AUC, calibration, nomogram)

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-pro-06
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   4. Core model evaluation (KM, forest, time‑dependent AUC, calibration, nomogram)
````

### 5 Robustness analysis (stability, ablation, sensitivity)

This section tests sensitivity to sampling, censoring, and feature changes. Robust conclusions should remain similar under reasonable perturbations.

````{r}
#| label: module4-pro-07
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   5. Robustness analysis (stability, ablation, sensitivity)
````

### 6 Decision curve analysis (DCA) and SHAP explanations

This section examines clinical utility and model explanations. Explanatory plots describe model behavior and do not establish causality.

````{r}
#| label: module4-pro-08
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   6. Decision curve analysis (DCA) and SHAP explanations
````

### 7 Ten advanced extensions (all executed sequentially)

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module4-pro-09
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   7. Ten advanced extensions (all executed sequentially)
````

### 8 Save final model and deployment (prediction + Shiny app)

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-pro-10
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   8. Save final model and deployment (prediction + Shiny app)
````

### 9 Final report summary

This section saves model objects, tables, and figures for reproducibility, review, and reuse.

````{r}
#| label: module4-pro-11
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   9. Final report summary
# =============================================================================

# ---------------------------------------------------------------------------
````

### 0 Environment and configuration

This section defines seeds, output paths, and dependencies. Update these paths and key parameters first when using new data.

````{r}
#| label: module4-pro-12
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 0. Environment and configuration
# ---------------------------------------------------------------------------
set.seed(2025)

# Version and output root
VERSION <- "Advanced_extensions"
OUTPUT_ROOT <- "./PrognosiX_Output"
OUTPUT_DIR <- file.path(OUTPUT_ROOT, VERSION)

# Helper to create subdirectories
sub_dir <- function(...) {
  d <- file.path(OUTPUT_DIR, ...)
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
  d
}

# Create version root and standard subfolders
sub_dir()
sub_dir("data")
sub_dir("01_feature_selection")
sub_dir("02_model")
sub_dir("03_evaluation")
sub_dir("04_robustness")
sub_dir("05_xai")

# Load required packages
library(icare)
required_pkgs <- c("mlr3", "mlr3proba", "mlr3viz", "mlr3learners", 
                   "survival", "survminer", "ggplot2", "caret", "dplyr",
                   "tidyr", "rms", "risksetROC", "survex", "dcurves",
                   "readxl", "patchwork", "ggrepel", "maxstat")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs)) install.packages(missing_pkgs)
invisible(lapply(required_pkgs, library, character.only = TRUE))

# Additional packages for extensions (checked inside each function)
ext_pkgs <- c("mlr3extralearners", "survcomp", "mlr3mbo", "bbotk", "MASS", "glmnet")
for (pkg in ext_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message("Note: Package '", pkg, "' not installed. Some advanced extensions may be skipped.")
  }
}

# ---------------------------------------------------------------------------
````

### 1 Data preparation (move time/status to info.data)

This section imports and organizes the data. Check dimensions and variable coding before continuing.

````{r}
#| label: module4-pro-13
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 1. Data preparation (move time/status to info.data)
# ---------------------------------------------------------------------------
raw_dis <- as.data.frame(read_excel("data/PMID37633276_DIA_plasmaproomic.xlsx", sheet = "Table S1B"))
raw_dis <- raw_dis[!grepl(x = raw_dis$`Case ID`, pattern = "^KC"), ]

# Extract numeric features (laboratory parameters)
mat <- raw_dis[, c("White blood cell count", 
                   "Lymphocytes%", "Monocytes%", "Neutrophilic granulocyte%", "Eosinophil%", 
                   "Basophil%", "Lymphocytes count", "Monocytes count", "Neutrophilic granulocyte count", 
                   "Eosinophil count", "Basophil count", "Red blood cell count", 
                   "Hemoglobin", "Hematocrit", "Mean red blood cell volume", "Mean cell hemoglobin", 
                   "Mean corpuscular hemoglobin concentration", "Red blood cell distribution width CV", 
                   "Red blood cell distribution width SD", "Platelet count", "Mean platelet volume", 
                   "PCT", "Platelet volume distribution width")]

index <- grepl(x = raw_dis$`Case ID`, pattern = "^HC")

# Clinical metadata (including time, status, and subgroup variables)
inf <- raw_dis[, c("Case ID", "Sample type", "Plasma proteome ID", "Tissue proteome ID", 
                   "Gender", "Age at diagnosis", "Tobacco smoking history", "Drinking history", 
                   "Vital status", "Days to last followup", "Days until death", 
                   "Combined days to last followup or death", "Progressive-free survival_Status", 
                   "Progressive-free survival_Months", "Tumor location", "Tumor size (cm)", 
                   "T-category", "Lymph node involvement", "Distant metastasis", 
                   "TNM Stage", "Grade", "Morphology")]

inf$group <- ifelse(index, '0', '1')
mat$group <- inf$group
mat <- as.matrix(mat)
storage.mode(mat) <- "numeric"

rownames(mat) <- inf$`Case ID`
rownames(inf) <- inf$`Case ID`
mat <- as.data.frame(mat)
mat$age <- inf$`Age at diagnosis`
mat$sex <- inf$Gender

inf$time <- inf$`Combined days to last followup or death`
inf$event <- inf$`Vital status`
inf$event <- ifelse(inf$event == "Dead", 1, 0)

# Keep only disease samples (group == 1)
inf <- inf[inf$group == 1, ]
mat <- mat[rownames(inf), ]
mat$time <- inf$time
mat$status <- inf$event

# Build Stat object with clean.data (features) and info.data (metadata including time/status)
stat_obj <- CreateStatObject(
  clean.data = mat,
  info.data = inf,
  group_col = "group",
  na.action = "allow"
)
cat("Stat object created.\n")

# Convert to PrognosiX: time/status are moved to info.data automatically
prog <- Stat_to_PrognosiX(stat_obj, "time", "status",
                          na_action = "omit",
                          min_events = 10,
                          verbose = TRUE)

# ---------------------------------------------------------------------------
````

### 2 Feature selection (ONLY on training set)

This section selects candidate features. Consider effect size, stability, and biological plausibility rather than a single significance measure.

````{r}
#| label: module4-pro-14
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 2. Feature selection (ONLY on training set)
# ---------------------------------------------------------------------------
task_full <- surv_extract_task(prog)
n <- task_full$nrow
train_idx <- caret::createDataPartition(task_full$data()[[task_full$target_names[2]]],
                                        p = 0.7, list = FALSE)
val_idx <- setdiff(seq_len(n), train_idx)

train_task <- task_full$clone()$filter(train_idx)
val_task <- task_full$clone()$filter(val_idx)

# Multi‑method feature selection: univariate Cox, LASSO, RF importance, Ridge, Elastic Net, VIMP
methods_to_run <- c("uni_cox", "lasso", "rf_imp", "ridge", "enet", "vimp")
feat_sel <- surv_feature_selection_multi(
  object = train_task,
  methods = methods_to_run,
  p_threshold = 0.1,
  top_ratio = 0.5,
  combine = "union",
  verbose = TRUE
)
selected_feats <- feat_sel$selected
cat("Selected features:", paste(selected_feats, collapse = ", "), "\n")
write.csv(feat_sel$method_table,
          file = file.path(sub_dir("01_feature_selection"), "feature_selection_methods.csv"),
          row.names = FALSE)

train_task <- train_task$select(selected_feats)
val_task <- val_task$select(selected_feats)

# Save train/validation data for reproducibility
train_data <- as.data.frame(train_task$data())
val_data <- as.data.frame(val_task$data())
write.csv(train_data, file = file.path(sub_dir("data"), "training_data.csv"), row.names = FALSE)
write.csv(val_data,   file = file.path(sub_dir("data"), "validation_data.csv"), row.names = FALSE)

# ---------------------------------------------------------------------------
````

### 3 Algorithm benchmarking and hyperparameter tuning

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-pro-15
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3. Algorithm benchmarking and hyperparameter tuning
# ---------------------------------------------------------------------------
# Compare Cox, LASSO, and Random Survival Forest (Ranger)
learners_list <- list(
  surv_get_learner("surv.coxph",     train_task),
  surv_get_learner("surv.cv_glmnet", train_task),
  surv_get_learner("surv.ranger",    train_task)
)
bmr1 <- surv_run_algorithm_benchmark(train_task, learners_list)
bmr2 <- surv_benchmark_learners(train_task, learner_ids = c("surv.coxph", "surv.ranger"), tune = FALSE)
bmr2_summary <- surv_summarize_benchmark(bmr2)
write.csv(bmr2_summary, file = file.path(sub_dir("02_model"), "algorithm_benchmark.csv"), row.names = FALSE)

# Select the best model (here Ranger) and perform hyperparameter tuning
best_id <- "surv.ranger"
tune <- surv_train_and_tune(train_task, best_id, tuning_budget = 25)
best_lrn <- tune$learner
cv_cindex <- tune$cv_performance
cat("CV C-index after tuning:", round(cv_cindex, 4), "\n")

# Train a Cox model for comparison (used in extensions)
cox_lrn <- surv_get_learner("surv.coxph", train_task)$train(train_task)

# ---------------------------------------------------------------------------
````

### 4 Core model evaluation

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-pro-16
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 4. Core model evaluation
# ---------------------------------------------------------------------------
# Warning: if no independent validation data is provided, some metrics are apparent
if (is.null(val_data)) {
  msg <- paste(
    "\n",
    "==================== WARNING ====================\n",
    "No external validation data (`val_data`) provided.\n",
    "KM curves, time‑dependent AUC, and nomogram will be\n",
    "computed on the TRAINING SET – these are apparent\n",
    "(optimistic) estimates. Do not report as generalizable.\n",
    "==================================================\n"
  )
  warning(msg, call. = FALSE, immediate. = TRUE)
  message(msg)
}

# Training set (apparent) performance
best_lrn$predict_type <- "distr"
train_perf <- surv_evaluate_model(best_lrn, train_task,
                                  measures = list(msr("surv.cindex"), msr("surv.graf")))
cat("Training apparent C-index:", round(train_perf$surv.cindex, 4), "\n")

# Validation set (unbiased) performance
val_pred <- best_lrn$predict(val_task)
val_cindex <- val_pred$score(msr("surv.cindex"))
cat("Validation C-index:", round(val_cindex, 4), "\n")

# Save performance metrics
perf_df <- data.frame(
  Dataset = c("Training", "Validation"),
  C_index = c(train_perf$surv.cindex, val_cindex)
)
write.csv(perf_df, file = file.path(sub_dir("03_evaluation"), "performance.csv"), row.names = FALSE)

# ---- 4.1 Risk stratification KM curves (multiple cutoff methods) ----
pred_train <- best_lrn$predict(train_task)
cut_df <- as.data.frame(train_task$data())
cut_df$risk <- pred_train$crank

# Compute optimal cutoff using surv_cutpoint; fallback to median if fails
cut_res <- tryCatch({
  surv_cutpoint(cut_df, time = "time", event = "status", variables = "risk")
}, error = function(e) {
  warning("surv_cutpoint failed: ", e$message, ". Using median split.")
  list(cutpoint = data.frame(cutpoint = median(cut_df$risk, na.rm = TRUE)))
})
opt_cutoff <- cut_res$cutpoint$cutpoint
cat("Optimal cutoff (from surv_cutpoint):", round(opt_cutoff, 4), "\n")

# Generate KM plots for four cutoff methods
# p_optimize and maxstat are exploratory; their p-values should not be used as confirmatory
km_median    <- surv_plot_risk_km(best_lrn, train_task, "median", risk_table = TRUE)
km_tertile   <- surv_plot_risk_km(best_lrn, train_task, "tertile")
km_quartile  <- surv_plot_risk_km(best_lrn, train_task, "quartile")
km_p_optimize <- surv_plot_risk_km(best_lrn, train_task, "p_optimize",
                                   n_boot = 100, fraction = 0.7)
km_maxstat   <- surv_plot_risk_km(best_lrn, train_task, "maxstat",
                                  n_boot = 100, minprop = 0.3)

# Save training KM plots

print(km_median)
print(km_tertile)
print(km_quartile)
print(km_p_optimize)
print(km_maxstat)

# Apply the training median cutoff to the validation set (fixed threshold)
train_cutoff <- get_cf(km_median)
if (is.null(train_cutoff)) train_cutoff <- opt_cutoff
km_val_fixed <- surv_plot_risk_km(best_lrn, val_task,
                                  cutoff_method = "custom",
                                  custom_cutoffs = train_cutoff,
                                  risk_table = TRUE)
print(km_val_fixed)

# ---- 4.2 Subgroup forest plot (using training set) ----
# Subset prog to training samples to ensure row consistency
train_prog <- prog
train_prog@clean.data <- prog@clean.data[train_idx, , drop = FALSE]
train_prog@info.data <- prog@info.data[train_idx, , drop = FALSE]
train_prog@survival.data <- prog@survival.data[train_idx, , drop = FALSE]

clinical_cols <- colnames(train_prog@info.data)
candidate_vars <- c("Gender", "Tobacco.smoking.history", "Drinking.history", "TNM.Stage")
subgroup_vars <- intersect(candidate_vars, clinical_cols)

forest <- surv_plot_subgroup_forest(
  learner = best_lrn,
  object = train_task,
  subgroup_vars = subgroup_vars,
  prog = train_prog,
  save_plot = TRUE,
  save_dir = sub_dir("03_evaluation"),
  file_name = "subgroup_forest.pdf"
)
print(forest)

# ---- 4.3 Time‑dependent AUC (training vs validation) ----
auc_train <- surv_plot_time_dependent_auc(best_lrn, train_task)
auc_val   <- surv_plot_time_dependent_auc(best_lrn, val_task)
auc_cmp   <- surv_plot_comparison_auc(best_lrn, train_task, val_task)
ggsave(file.path(sub_dir("03_evaluation"), "AUC_comparison.pdf"), auc_cmp, width = 8, height = 5)

# ---- 4.4 Calibration curves (training vs validation) ----
cal_train <- surv_plot_calibration(best_lrn, train_task, time_point = 200)
cal_cmp   <- surv_plot_comparison_calibration(best_lrn, train_task, val_task, time_point = 200)
ggsave(file.path(sub_dir("03_evaluation"), "Calibration_train.pdf"), cal_train, width = 8, height = 6)
ggsave(file.path(sub_dir("03_evaluation"), "Calibration_comparison.pdf"), cal_cmp, width = 8, height = 6)

# ---- 4.5 External validation (using the validation set) ----
val_data_ext <- as.data.frame(task_full$data())[val_idx, selected_feats, drop = FALSE]
val_data_ext$time <- task_full$data()[[task_full$target_names[1]]][val_idx]
val_data_ext$status <- task_full$data()[[task_full$target_names[2]]][val_idx]
val_res <- surv_predict_on_validation(best_lrn, val_data_ext, train_task)
val_cindex_alt <- val_res$prediction$score(msr("surv.cindex"))
cat("Validation C-index (alt):", round(val_cindex_alt, 4), "\n")

# ---- 4.6 Nomogram (with proportional hazards test) ----
nom <- surv_generate_nomogram(train_task,
                              selected_features = head(selected_feats, 5),
                              time_points = c(90, 180, 365),
                              time_unit = "days")
# The nomogram plot is drawn; PH test results are printed to console.

# ---------------------------------------------------------------------------
````

### 5 Robustness analysis

This section tests sensitivity to sampling, censoring, and feature changes. Robust conclusions should remain similar under reasonable perturbations.

````{r}
#| label: module4-pro-17
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 5. Robustness analysis
# ---------------------------------------------------------------------------
````

### 5.1 Feature stability (LASSO bootstrapping)

This section tests sensitivity to sampling, censoring, and feature changes. Robust conclusions should remain similar under reasonable perturbations.

````{r}
#| label: module4-pro-18
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 5.1 Feature stability (LASSO bootstrapping)
stab <- surv_analyze_feature_stability(train_task, "time", "status", n_repeat = 20, alpha = 1)
cat("Stability index:", round(stab$stability_index, 4), "\n")
ggsave(file.path(sub_dir("04_robustness"), "stability_plot.pdf"), stab$plot, width = 8, height = 6)
write.csv(stab$frequencies, file.path(sub_dir("04_robustness"), "stability_frequencies.csv"), row.names = FALSE)

````

### 5.2 Feature ablation (drop each feature and measure C-index drop)

This section tests sensitivity to sampling, censoring, and feature changes. Robust conclusions should remain similar under reasonable perturbations.

````{r}
#| label: module4-pro-19
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 5.2 Feature ablation (drop each feature and measure C-index drop)
abl <- surv_analyze_feature_ablation(train_task, best_id, selected_feats)
ggsave(file.path(sub_dir("04_robustness"), "ablation_plot.pdf"), abl$plot, width = 8, height = 6)
write.csv(abl$results, file.path(sub_dir("04_robustness"), "ablation_results.csv"), row.names = FALSE)

````

### 5.3 Model sensitivity (sample size and censoring rate)

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-pro-20
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 5.3 Model sensitivity (sample size and censoring rate)
sens_sample <- surv_analyze_model_sensitivity(train_task, best_id, analysis_type = "sample_size")
sens_censor <- surv_analyze_model_sensitivity(train_task, best_id, analysis_type = "censoring")
ggsave(file.path(sub_dir("04_robustness"), "sensitivity_sample.pdf"), sens_sample$plot, width = 8, height = 5)
ggsave(file.path(sub_dir("04_robustness"), "sensitivity_censor.pdf"), sens_censor$plot, width = 8, height = 5)
write.csv(sens_sample$results, file.path(sub_dir("04_robustness"), "sensitivity_sample.csv"), row.names = FALSE)
write.csv(sens_censor$results, file.path(sub_dir("04_robustness"), "sensitivity_censor.csv"), row.names = FALSE)

# ---------------------------------------------------------------------------
````

### 6 Decision curve analysis (DCA) and SHAP explanations (on validation set)

This section examines clinical utility and model explanations. Explanatory plots describe model behavior and do not establish causality.

````{r}
#| label: module4-pro-21
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 6. Decision curve analysis (DCA) and SHAP explanations (on validation set)
# ---------------------------------------------------------------------------
# Determine a valid evaluation time from validation event times
val_time <- val_task$data()[[val_task$target_names[1]]]
val_status <- val_task$data()[[val_task$target_names[2]]]
event_times <- val_time[val_status == 1]
if (length(event_times) == 0) stop("No events in validation set.")
eval_time <- round(quantile(event_times, 0.5, na.rm = TRUE), 0)
if (eval_time < min(event_times)) eval_time <- min(event_times)
cat("Using eval_time =", eval_time, "for DCA\n")

# DCA
dca_res <- plot_dca_survival(
  learners = list("Tuned" = best_lrn),
  object = val_task,
  eval_time = eval_time,
  clin_range = c(0.05, 0.5),
  print_stats = TRUE
)
ggsave(file.path(sub_dir("05_xai"), "dca_plot.pdf"), dca_res$plot, width = 7, height = 5)
write.csv(dca_res$table, file.path(sub_dir("05_xai"), "dca_table.csv"), row.names = FALSE)

# Global SHAP (SurvSHAP(t))
shap_global <- surv_explain_shap(
  learner = best_lrn,
  task = val_task,
  type = "global",
  n_explain = 30,
  n_background = 10,
  n_timepoints = 10,
  n_top_features = 6,
  verbose = TRUE
)
ggsave(file.path(sub_dir("05_xai"), "shap_global_bar.pdf"), shap_global$plots$bar_plot, width = 8, height = 6)
if (!is.null(shap_global$plots$line_plot)) {
  ggsave(file.path(sub_dir("05_xai"), "shap_global_line.pdf"), shap_global$plots$line_plot, width = 8, height = 6)
}
write.csv(shap_global$shap_long, file.path(sub_dir("05_xai"), "shap_values_global.csv"), row.names = FALSE)

# Beeswarm plot
beeswarm_plot <- surv_plot_shap_beeswarm(shap_global, top_n = 6, method = "beeswarm",
                                         title = "Global SurvSHAP (validation set)")
ggsave(file.path(sub_dir("05_xai"), "shap_beeswarm.pdf"), beeswarm_plot, width = 9, height = 6)

# =============================================================================
````

### 7 Ten advanced extensions (all executed sequentially)

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module4-pro-22
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 7. Ten advanced extensions (all executed sequentially)
# =============================================================================
cat("\n========== Running 10 advanced extensions ==========\n")

# ---- 7.1 Competing risks (Fine‑Gray or cause‑specific Cox) ----
# Purpose: Correctly estimate cumulative incidence when competing events exist.
# Method: Fine‑Gray via mlr3extralearners or cause‑specific Cox.
cat("\n--- 7.1 Competing risks (Fine‑Gray / cause‑specific Cox) ---\n")
if (requireNamespace("mlr3extralearners", quietly = TRUE)) {
  fg_res <- tryCatch({
    run_competing_risks(prog, model_type = "finegray", cause = 1, tuning_budget = 20)
  }, error = function(e) {
    cat("Competing risks failed:", e$message, "\n")
    NULL
  })
  if (!is.null(fg_res)) {
    cat("Fine‑Gray CV C-index:", round(fg_res$cv_cindex, 4), "\n")
    saveRDS(fg_res, file.path(sub_dir("05_xai"), "competing_risks.rds"))
  }
} else {
  cat("Skipping (mlr3extralearners not installed).\n")
}

# ---- 7.2 Time‑dependent ROC comparison (Ranger vs Cox) ----
# Purpose: Compare AUC trajectories and obtain bootstrap confidence intervals.
# Method: timeROC with bootstrapping.
cat("\n--- 7.2 Time‑dependent ROC comparison (Ranger vs Cox) ---\n")
td_roc <- tryCatch({
  plot_tdROC_comparison(best_lrn, cox_lrn, val_task, n_boot = 50, save_plot = TRUE)
}, error = function(e) {
  cat("tdROC comparison failed:", e$message, "\n")
  NULL
})

# ---- 7.3 Multi‑time calibration curves ----
# Purpose: Assess calibration at multiple clinically relevant horizons.
# Method: Separate calibration plots at each time point, combined with patchwork.
cat("\n--- 7.3 Multi‑time calibration curves (90, 180, 365 days) ---\n")
tryCatch({
  plot_multitime_calibration(best_lrn, val_task, time_points = c(90, 180, 365), save_plot = TRUE)
}, error = function(e) {
  cat("Multi‑time calibration failed:", e$message, "\n")
})

# ---- 7.4 Repeated cross‑validation (with parallel support) ----
# Purpose: Obtain robust performance estimates with confidence intervals.
# Method: 5 repeats of 5‑fold CV, compute Bootstrap CI for C‑index.
cat("\n--- 7.4 Repeated cross‑validation (5×5‑fold CV) ---\n")
cv_res <- tryCatch({
  repeated_cv_evaluation(prog, "surv.ranger", n_repeats = 5, folds = 5, parallel = FALSE)
}, error = function(e) {
  cat("Repeated CV failed:", e$message, "\n")
  NULL
})
if (!is.null(cv_res)) {
  cat("Repeated CV summary:\n")
  print(cv_res$summary)
  write.csv(cv_res$results, file.path(sub_dir("04_robustness"), "repeated_cv_results.csv"), row.names = FALSE)
}

# ---- 7.5 DeLong test for comparing C‑indices ----
# Purpose: Statistically compare discriminative ability of Ranger vs Cox.
# Method: survcomp::cindex.comp (bootstrap‑based DeLong).
cat("\n--- 7.5 DeLong test (Ranger vs Cox) ---\n")
delong <- tryCatch({
  compare_models_delong(best_lrn, cox_lrn, val_task, n_boot = 500)
}, error = function(e) {
  cat("DeLong test failed:", e$message, "\n")
  NULL
})
if (!is.null(delong)) {
  cat("DeLong p‑value:", delong$p.value, "\n")
  cat("C‑index Ranger:", round(delong$C_index_model1, 4), "\n")
  cat("C‑index Cox:    ", round(delong$C_index_model2, 4), "\n")
  write.csv(data.frame(
    Model1_Cindex = delong$C_index_model1,
    Model2_Cindex = delong$C_index_model2,
    Difference = delong$difference,
    P_value = delong$p.value,
    CI_lower = delong$ci_lower,
    CI_upper = delong$ci_upper
  ), file.path(sub_dir("05_xai"), "delong_test.csv"), row.names = FALSE)
}

# ---- 7.6 Bayesian hyperparameter optimisation ----
# Purpose: More efficient search than random search, with automatic fallback.
# Method: mlr3mbo, falls back to random search on error.
cat("\n--- 7.6 Bayesian optimisation (Ranger) ---\n")
bayes_tune <- tryCatch({
  tune_with_bayes(train_task, "surv.ranger", tuning_budget = 30)
}, error = function(e) {
  cat("Bayesian optimisation failed:", e$message, "\n")
  NULL
})
if (!is.null(bayes_tune)) {
  saveRDS(bayes_tune, file.path(sub_dir("02_model"), "bayes_tune.rds"))
  cat("Bayesian tuned C‑index:", round(bayes_tune$cv_performance, 4), "\n")
}

# ---- 7.7 AIC stepwise variable selection ----
# Purpose: Complementary traditional feature selection using AIC.
# Method: MASS::stepAIC on a Cox model (bidirectional).
cat("\n--- 7.7 AIC stepwise selection (both directions) ---\n")
step <- tryCatch({
  stepwise_variable_selection(prog, direction = "both")
}, error = function(e) {
  cat("Stepwise selection failed:", e$message, "\n")
  NULL
})
if (!is.null(step)) {
  cat("Stepwise selected features:", paste(step$selected_features, collapse = ", "), "\n")
  write.csv(data.frame(Features = step$selected_features),
            file.path(sub_dir("01_feature_selection"), "stepwise_features.csv"), row.names = FALSE)
}

# ---- 7.8 Calibration‑in‑the‑large ----
# Purpose: Check overall calibration (mean predicted vs observed event rate).
# Method: Logistic regression of observed binary outcome on logit(predicted probability).
cat("\n--- 7.8 Calibration‑in‑the‑large (t = 365 days) ---\n")
cal_large <- tryCatch({
  calibration_in_the_large(best_lrn, val_task, time_point = 365)
}, error = function(e) {
  cat("Calibration‑in‑the‑large failed:", e$message, "\n")
  NULL
})
if (!is.null(cal_large)) {
  cat("Calibration intercept:", round(cal_large$intercept, 4), "\n")
  cat("Calibration slope:    ", round(cal_large$slope, 4), "\n")
  cat("Mean predicted:       ", round(cal_large$mean_predicted, 4), "\n")
  cat("Observed:             ", round(cal_large$observed, 4), "\n")
  write.csv(data.frame(
    Intercept = cal_large$intercept,
    Slope = cal_large$slope,
    Mean_Predicted = cal_large$mean_predicted,
    Observed = cal_large$observed,
    N_at_risk = cal_large$n_at_risk
  ), file.path(sub_dir("03_evaluation"), "calibration_large.csv"), row.names = FALSE)
}

# ---- 7.9 External validation (Hosmer‑Lemeshow test) ----
# Purpose: Assess calibration goodness‑of‑fit on validation set.
# Method: Group patients by predicted risk, compare observed vs expected events, χ² test.
cat("\n--- 7.9 Survival Hosmer‑Lemeshow test (t = 365 days, 10 groups) ---\n")
hl <- tryCatch({
  external_validation_test(best_lrn, val_task, time_point = 365, n_groups = 10)
}, error = function(e) {
  cat("HL test failed:", e$message, "\n")
  NULL
})
if (!is.null(hl)) {
  cat("HL chi‑square:", round(hl$chi_square, 4), "df:", hl$df, "p‑value:", round(hl$p_value, 4), "\n")
  write.csv(data.frame(
    Chi_square = hl$chi_square,
    df = hl$df,
    P_value = hl$p_value,
    N_total = hl$n_total
  ), file.path(sub_dir("03_evaluation"), "hl_test.csv"), row.names = FALSE)
}

# ---- 7.10 Clinical impact curve ----
# Purpose: Show number of high‑risk patients and events across thresholds.
# Method: Bootstrap confidence bands for both curves.
cat("\n--- 7.10 Clinical impact curve (t = 365 days) ---\n")
tryCatch({
  plot_clinical_impact(best_lrn, val_task, time_point = 365, n_boot = 100, save_plot = TRUE)
}, error = function(e) {
  cat("Clinical impact curve failed:", e$message, "\n")
})

cat("\n========== All advanced extensions completed ==========\n")

# ---------------------------------------------------------------------------
````

### 8 Save final PrognosiX object and deploy

This section applies the trained object to new samples. Validate input format, performance, and target population in an independent cohort before deployment.

````{r}
#| label: module4-pro-23
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 8. Save final PrognosiX object and deploy
# ---------------------------------------------------------------------------
prog@best.model <- list(
  learner_id = best_id,
  learner = best_lrn,
  best_params = tune$best_params,
  cv_cindex = cv_cindex,
  features = selected_feats,
  cutoff = opt_cutoff,
  decision_type = "binary",
  train_cols = train_task$feature_names
)
prog@subgroup.risk <- list(
  benchmark_table = bmr1$table,
  stability = stab,
  ablation = abl,
  validation = list(cindex = val_cindex, n = val_task$nrow)
)
prog@split.data <- list(train_idx = train_idx, val_idx = val_idx)
saveRDS(prog, file = file.path(sub_dir("data"), "prog_obj_final.rds"))

# ---- 8.1 Predict risk for new patients ----
#new_patients <- data.frame(
#  Mean_corpuscular_hemoglobin_concentration = c(33, 34),
#  Basophil_count = c(0.02, 0.03),
#  Monocytes = c(0.6, 0.7),
#  Eosinophil_count=c(0.02, 0.03),
#  Basophil=c(0.6, 0.7),
#  Neutrophilic_granulocyte = c(4.2, 5.1),
#  Red_blood_cell_distribution_width_CV = c(12.3, 11.8),
#  Mean_red_blood_cell_volume = c(89, 92),
#  Lymphocytes_count = c(1.8, 2.1),
#  Platelet_volume_distribution_width = c(12.0, 13.1),
#  Lymphocytes = c(1.8, 2.1),
#  Monocytes_count = c(0.6, 0.7),
#  Neutrophilic_granulocyte_count = c(4.2, 5.1),
#  Mean_platelet_volume = c(10.5, 11.0),
#  row.names = c("patient_1", "patient_2")
#)

#risk_scores <- predict_prognosix(prog, new_patients, impute = TRUE)
#pred_groups_med <- predict_risk_groups(prog, new_patients, cutoff_method = "median")
#pred_groups_cus <- predict_risk_groups(prog, new_patients,
#                                       cutoff_method = "custom",
#                                       custom_cutoffs = opt_cutoff)
#cat("Risk scores:\n"); print(risk_scores)
#cat("Median groups:\n"); print(pred_groups_med)
#cat("Custom groups:\n"); print(pred_groups_cus)

# ---- 8.2 Deployment manager and Shiny app ----
manager <- New_Prog_Manager(prog)

if (interactive()) {
  use_app_theme_grey()
  set_prog_app_text(
    title = "Survival Risk Predictor",
    citation_text = "Data: PMID37633276_DIA_plasmaproomic cohort."
  )
  set_prog_app_theme(
    primary_color = "#2c7fb8", background_color = "#f0f4f8", sidebar_color = "#e9ecef",
    box_background = "#ffffff", label_color = "#2c7fb8",
    run_button_gradient_start = "#2c7fb8", run_button_gradient_end = "#1d4e6e",
    risk_high_color = "#d9534f", risk_medium_color = "#f0ad4e", risk_low_color = "#5cb85c",
    table_header_color = "#2c7fb8", font_family = "Arial, sans-serif", font_size_base = 14
  )
  launch_prog_deploy_app(
    manager,
    var_dict = data.frame(
      Feature = selected_feats,
      Description = paste("Predictor", seq_along(selected_feats)),
      Units = rep("--", length(selected_feats))
    ),
    project_info = list(
      abstract = paste("Prognostic model trained on", train_task$nrow, "patients."),
      citation = "PMID37633276_DIA_plasmaproomic cohort."
    )
  )
}

# ---------------------------------------------------------------------------
````

### 9 Final report

This section saves model objects, tables, and figures for reproducibility, review, and reuse.

````{r}
#| label: module4-pro-24
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 9. Final report
# ---------------------------------------------------------------------------
cat("\n========================================\n")
cat("Module 4 (advanced, leakage‑free) finished successfully!\n")
cat("Best model:", best_id,
    "| CV C-index (training):", round(cv_cindex, 4),
    "| Validation C-index:", round(val_cindex, 4), "\n")
cat("Optimal cutoff:", round(opt_cutoff, 4), "\n")
cat("All results saved under:", OUTPUT_DIR, "\n")
cat("========================================\n")
````
