# Module 2: Classification Modeling — Advanced

The advanced `Train_Model` workflow covers algorithm availability checks, RFE/GA/SA and built-in feature selection, preprocessing benchmarks, model comparison, ensembles, hyperparameter tuning, SHAP and DALEX explanations, clinically interpretable thresholds, NRI/IDI, deployment, and an optional Shiny interface.

## Key Configuration

Start by setting `INPUT_STAT_OBJ`, the outcome definition, and `OUTPUT_DIR`. Optional packages are loaded only by the methods that need them. Treat apparent performance from feature selection, tuning, and model comparison as development evidence until it is confirmed in an untouched cohort; all figures and tables are saved in a structured output tree for that audit.

## Step-by-step Workflow

<!-- split-into-step-chunks -->

The workflow is divided into short stages. Each note explains the purpose, key checks, and interpretation limits. Generated plots and tables appear directly below their code chunk.

### Environment and Global Setup

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module2-advanced-01
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# =============================================================================
# icare Package Vignette — Module 2: Modeling Pipeline
# ADVANCED / PUBLICATION EDITION
# =============================================================================
# Audience : Advanced users who want to see (and customize) every stage of
#            building, tuning, explaining, and clinically validating a model.
# Input    : "stat_obj.rds", produced by Module 1 (data cleaning).
#
# Pipeline covered in this script:
````

### 1 Convert the cleaned Stat object into a modeling object

This section creates the icare analysis object that stores data, metadata, and workflow state for all later steps.

````{r}
#| label: module2-advanced-02
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   1) Convert the cleaned Stat object into a modeling object
````

### 2 Inspect candidate algorithms and their hyperparameters

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-03
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   2) Inspect candidate algorithms and their hyperparameters
````

### 3 Feature selection (built-in importance, then a multi-method

This section selects candidate features. Consider effect size, stability, and biological plausibility rather than a single significance measure.

````{r}
#| label: module2-advanced-04
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   3) Feature selection (built-in importance, then a multi-method
#      RFE/GA/SA pipeline, then a sensitivity/elbow-based elimination)
````

### 4 Preprocessing benchmark, train/test split & scaling

This section preprocesses the data. Review missingness, outlier handling, and normalization before accepting the defaults.

````{r}
#| label: module2-advanced-05
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   4) Preprocessing benchmark, train/test split & scaling
````

### 5 Train several algorithms and compare them

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-06
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   5) Train several algorithms and compare them
````

### 6 Ensemble models (stacking / averaging / weighting / voting)

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-07
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   6) Ensemble models (stacking / averaging / weighting / voting)
````

### 7 Hyperparameter tuning of the best model

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-08
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   7) Hyperparameter tuning of the best model
````

### 8 Explainability (SHAP, break-down, partial dependence, etc.)

This section examines clinical utility and model explanations. Explanatory plots describe model behavior and do not establish causality.

````{r}
#| label: module2-advanced-09
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   8) Explainability (SHAP, break-down, partial dependence, etc.)
````

### 9 Clinical layer: subgroup analysis, confounder adjustment,

This section examines clinical utility and model explanations. Explanatory plots describe model behavior and do not establish causality.

````{r}
#| label: module2-advanced-10
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   9) Clinical layer: subgroup analysis, confounder adjustment,
#      decision thresholds, NRI/IDI model comparison
````

### 10 Deployment: a prediction function and an optional Shiny app

This section applies the trained object to new samples. Validate input format, performance, and target population in an independent cohort before deployment.

````{r}
#| label: module2-advanced-11
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#  10) Deployment: a prediction function and an optional Shiny app
#
# Every `save_dir` below points into a sub-folder of OUTPUT_DIR so all
# figures/tables from a run are easy to locate.
# =============================================================================

rm(list = ls())
library(icare)
library(caret)
#devtools::document("../../icare-git/")
#devtools::install("../../icare-git/")
# ---------------------------------------------------------------------------
````

### 0 CONFIG

This section defines seeds, output paths, and dependencies. Update these paths and key parameters first when using new data.

````{r}
#| label: module2-advanced-12
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 0. CONFIG
# ---------------------------------------------------------------------------
INPUT_STAT_OBJ <- "stat_obj.rds"                 # produced by Module 1
OUTPUT_DIR     <- "./Module2_Advanced_Output"
SEED           <- 123

dir.create(OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)
sub_dir <- function(...) {
  d <- file.path(OUTPUT_DIR, ...)
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
  d
}

# =============================================================================
````

### 1 Load the Stat Object and Convert it to a Modeling Object

This section imports and organizes the data. Check dimensions and variable coding before continuing.

````{r}
#| label: module2-advanced-13
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 1. Load the Stat Object and Convert it to a Modeling Object
# =============================================================================
stat_obj <- readRDS(INPUT_STAT_OBJ)

# ConvertObject() re-wraps the cleaned data (`clean.data`) into a
# `Train_Model` object, icare's container for the modeling stage.
model_obj <- ConvertObject(stat_obj, to = "Train_Model")
InspectObject(model_obj)

# The outcome column must be an explicit factor for classification models.
model_obj@clean.df[[model_obj@group_col]] <-
  factor(model_obj@clean.df[[model_obj@group_col]])
cat("Outcome distribution:\n")
print(table(model_obj@clean.df[[model_obj@group_col]]))

set.seed(SEED)
idx <- createDataPartition(model_obj@clean.df[[model_obj@group_col]], p = 0.7, list = FALSE)
model_obj@split.data <- list(
  training = model_obj@clean.df[idx, ],
  testing  = model_obj@clean.df[-idx, ]
)
model_obj_train <- CreateModelObject(
  data      = model_obj@split.data$training,
  group_col = model_obj@group_col
)

InspectObject(model_obj_train)
# =============================================================================
````

### 2 Algorithm Inspection

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-14
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 2. Algorithm Inspection
# =============================================================================
# Before picking algorithms, it helps to see what hyperparameters each one
# exposes (via caret) and which ones support built-in variable importance.
# Load data
data(allmodel)
df1=allmodel
# List all presets
df2=list_presets()
# Get all linear models
models_of_interest <- get_models(preset = "core")

cat("\n--- Hyperparameters for candidate models ---\n")
for (m in models_of_interest) {
  info <- caret::getModelInfo(m, regex = FALSE)[[1]]
  cat("\n", m, ":", paste(info$parameters$parameter, collapse = ", "))
}

cat("\n\n--- Built-in variable-importance support ---\n")
print(check_varImp_availability(models_of_interest))

# =============================================================================
````

### 3 Feature Selection

This section selects candidate features. Consider effect size, stability, and biological plausibility rather than a single significance measure.

````{r}
#| label: module2-advanced-15
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3. Feature Selection
# =============================================================================
fs_dir <- sub_dir("01_feature_selection")

## 3.1 Quick pass: built-in importance from a couple of fast models.
## Models that don't support importance are skipped automatically.
builtin <- FeatureSelectBuiltin(
  object = model_obj_train,
  models = models_of_interest,
  top_n  = 15,
  seed   = SEED
)
print(builtin$importance_table)
PlotBuiltinImportance(builtin_result = builtin, save_dir = fs_dir, save_plot = TRUE)

## 3.2 Thorough pass: combine Recursive Feature Elimination (RFE), a Genetic
## Algorithm (GA), and Simulated Annealing (SA), then take the UNION of
## features any method selected. This is slower but more robust than any
## single method — useful when you want a defensible feature set for a paper.
fs <- FeatureSelectionPipeline(
  object    = model_obj_train,
  methods   = c("ga", "rfe", "sa"),
  combine   = "union",
  rfe_args  = list(method = "cv", number = 2, sizes = c(5, 10, 20)),
  sa_args   = list(iters = 2, method = "cv", number = 5, repeats = 1, improve = 3),
  ga_args   = list(ga_func = caret::rfGA, iters = 2, popSize = 5,
                    method = "cv", number = 5,
                    allowParallel = TRUE, genParallel = TRUE),
  upset_plot = TRUE,
  save_plot  = TRUE,
  save_dir   = fs_dir,
  seed       = SEED
)

## 3.3 Diagnostic plots for the multi-method feature-selection run.
PlotRFE(fs$results$RFE, metric = "Accuracy", show_optimal = TRUE,y_limits = NULL,
        save_plot = TRUE, save_dir = fs_dir)
PlotRFEImportance(fs$results$RFE, top_n = 20, save_plot = TRUE, save_dir = fs_dir)
PlotGA(fs$results$GA, metric = "Accuracy", save_plot = TRUE, save_dir = fs_dir)
PlotGAFrequency(fs$results$GA, top_n = 30, save_plot = TRUE, save_dir = fs_dir)
PlotFeatureConsensus(fs, show_counts = TRUE, save_plot = TRUE, save_dir = fs_dir)
PlotFeatureStability(fs, top_n = 30, save_plot = TRUE, save_dir = fs_dir,palette_name = 'BottleRocket2')
PlotFeatureComparison(fs, metric = "Accuracy", save_plot = TRUE, save_dir = fs_dir)

## 3.4 A complementary, sensitivity-based approach: train models against an
## increasing number of features, then use an "elbow" rule to find the point
## where adding more features stops helping (either an absolute performance
## tolerance, or a stability window).
elim <- run_feature_elimination(model_obj_train, models = models_of_interest,
                                 number = 5, smooth_span = 0.2)
res_perf   <- select_elbow(elim, "perf_tolerance", tol = 0.05)

plot_elbow(elim, best_features = res_perf$best_features, ci_style = "ribbon",
           save_plot = TRUE, save_dir = fs_dir)

features_perf <- get_selected_features(elim, "perf_tolerance", tol = 0.05)
all_methods   <- get_selected_features(elim, methods = c("perf_tolerance"),
                                        tol = 0.05, window_size = 40, stability_tol = 0.005)

## 3.5 Commit to a final feature set. Here we use the random-forest result
## from the performance-tolerance elbow; swap in `builtin`, `fs$selected`,
## or `all_methods` if you prefer a different method's result.
model_obj_train<- ApplyFeatureSelection(model_obj_train, features_perf$perf_tolerance$best_features$glmnet)
model_obj<- ApplyFeatureSelection(model_obj, features_perf$perf_tolerance$best_features$glmnet)
# =============================================================================
````

### 4 Preprocessing Benchmark

This section preprocesses the data. Review missingness, outlier handling, and normalization before accepting the defaults.

````{r}
#| label: module2-advanced-16
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 4. Preprocessing Benchmark
# =============================================================================
# Before committing to one imputation/normalization recipe, compare a grid of
# (algorithm x imputation x normalization) combinations by cross-validated
# performance. This is optional but strengthens a methods section.
bench <- LogisticDiagnosticBenchmark(
  object = model_obj_train,
  preProcess = c("center", "scale"),
  smote = TRUE,save_plot = TRUE, save_dir = fs_dir)

#algorithms_list=c( 'glm',"lda", "knn","nb","rpart","rf","svmRadial","xgbTree")
algorithms_list=models_of_interest
bench_result <- PreprocessingBenchmark(
  data            = model_obj_train,
  group_col       = model_obj_train@group_col,
  algorithms      = algorithms_list,
  impute_methods  = c("none", "median", "knn", "bag"),
  norm_methods    = list("none", c("center"), c("center", "scale"))
)

print(bench_result)
PlotBenchmarkForest(bench_result, save_plot = TRUE,  save_dir  =fs_dir,global_median = F)
PlotBenchmarkForest(bench_result, save_plot = TRUE,  save_dir  =fs_dir,global_median = F,metric = 'Accuracy')
write.csv(bench_result, file.path(fs_dir, "preprocessing_benchmark.csv"), row.names = FALSE)
# =============================================================================
````

### 5 Train / Test Split & Scaling

This section creates training and validation sets. Restrict feature selection and parameter estimation to the training set to prevent leakage.

````{r}
#| label: module2-advanced-17
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 5. Train / Test Split & Scaling
# =============================================================================
preProc <- preProcess(
  model_obj@split.data$training[, setdiff(colnames(model_obj@split.data$training), model_obj@group_col)],
  method = c("center", "scale", "YeoJohnson", "bagImpute")
)
model_obj@split.scale.data <- list(
  training = predict(preProc, model_obj@split.data$training),
  testing  = predict(preProc, model_obj@split.data$testing)
)
model_obj@filtered.set <- model_obj@split.scale.data

# =============================================================================
````

### 6 Model Training

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-18
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 6. Model Training
# =============================================================================
model_dir <- sub_dir("02_models")

model_obj <- ModelTrainAnalysis(
  object     = model_obj,metric_selection = 'ROC',
  preProcess = NULL,
  tuneLength = 3,
  methods    = models_of_interest,
  control    = list(method = "repeatedcv", number = 5, repeats = 2),
  imbalance_handling = "auto", 
  imbalance_threshold = 0.2,
  save_plots = TRUE,
  save_dir   = model_dir,
  seed       = SEED
)

best_row <- model_obj@all.results[which.max(model_obj@all.results$auc), ]
cat("Best model by AUC:", best_row$Model, "\n")
model_obj <- SelectBestModel(model_obj, metric = "auc")

# =============================================================================
````

### 7 Model Comparison Visualizations

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-19
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 7. Model Comparison Visualizations
# =============================================================================
PlotTopModelAUC(model_obj, top_n = 4, save_plot = TRUE, save_dir = model_dir)
PlotProbDensity(model_obj, save_plot = TRUE, save_dir = model_dir)
PlotModelComparison(
  model_obj, top_n = 4,
  metrics = c("auc", "Sensitivity", "Specificity", "accuracy_score", "f1_score"),
  palette_name = "Darjeeling1", save_plot = TRUE, save_dir = model_dir,
  width = 12, height = 5
)

PlotModelParallel(
  model_obj, top_n = 4,
  metrics = c("auc", "Sensitivity", "Specificity", "accuracy_score", "f1_score"),
  save_plot = TRUE, save_dir = model_dir
)
PlotMultiROC(model_obj, test_data = model_obj@split.scale.data$testing,
             palette_name = "Darjeeling1", save_plot = TRUE, save_dir = model_dir)

# Pick a trained model to inspect. "rf" is included in the model list above.
PlotConfusionMatrix(model_obj, test_data = model_obj@split.scale.data$testing,
                    model_name = "rf", threshold = 0.6, prevalence = 0.3,
                    save_plot = TRUE, save_dir = model_dir)
PlotFeatureImportance(model_obj, top_n = 20, save_plot = TRUE, save_dir = model_dir)
PlotCalibration(model_obj, model_name = "rf",
                test_data = model_obj@filtered.set$training,
                save_plot = TRUE, save_dir = model_dir)
PlotCalibration(model_obj, model_name = "rf",
                test_data = model_obj@filtered.set$testing, 
                save_plot = TRUE, save_dir = model_dir)
# =============================================================================
````

### 8 Ensemble Models

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-20
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 8. Ensemble Models
# =============================================================================
# Train several ensembling strategies on independent copies of model_obj so
# they don't overwrite each other, then compare their held-out AUC.
ens_dir <- sub_dir("03_ensemble")
model_avg    <- TrainEnsemble(model_obj, strategy = "average",  top_n = 4)
model_stack  <- tryCatch(
  TrainEnsemble(model_obj, strategy = "stacking", meta_method = "glm", top_n = 4),
  error = function(e) { message("Stacking unavailable; using average ensemble: ", e$message); model_avg }
)
model_wgt    <- tryCatch(
  TrainEnsemble(model_obj, strategy = "weighted", top_n = 4),
  error = function(e) { message("Weighted ensemble unavailable; using average ensemble: ", e$message); model_avg }
)
model_vote   <- TrainEnsemble(model_obj, strategy = "voting",   top_n = 4)

# A manually-specified weighting is also supported, e.g. if domain knowledge
# suggests one algorithm should count more than another.
custom_wts   <- c(glm = 0.1, rf = 0.5, gbm = 0.3, svmRadial = 0.1)
model_custom <- tryCatch(
  TrainEnsemble(model_obj, strategy = "weighted", weights = custom_wts, top_n = 4),
  error = function(e) { message("Custom weighting unavailable; using average ensemble: ", e$message); model_avg }
)

test_data <- predict(preProc, model_obj@clean.df)
probs_stack  <- PredictEnsemble(model_stack,  test_data)
probs_avg    <- PredictEnsemble(model_avg,    test_data)
probs_wgt    <- PredictEnsemble(model_wgt,    test_data)
probs_custom <- PredictEnsemble(model_custom, test_data)

library(pROC)
ensemble_auc <- data.frame(
  Strategy = c("Stacking (or Average fallback)", "Average", "Weighted (or Average fallback)", "CustomWeighted (or Average fallback)"),
  AUC = c(
    as.numeric(auc(roc(test_data[[model_obj@group_col]], probs_stack))),
    as.numeric(auc(roc(test_data[[model_obj@group_col]], probs_avg))),
    as.numeric(auc(roc(test_data[[model_obj@group_col]], probs_wgt))),
    as.numeric(auc(roc(test_data[[model_obj@group_col]], probs_custom)))
  )
)
print(ensemble_auc)
write.csv(ensemble_auc, file.path(ens_dir, "ensemble_auc_comparison.csv"), row.names = FALSE)

# Voting returns hard class labels rather than probabilities, so it is
# compared separately by accuracy if needed — omitted here for brevity.

model_obj <- SelectBestModel(model_obj, metric = "auc")
cat("Best single model:", model_obj@best.model.result$model_type, "\n")

# =============================================================================
````

### 9 Hyperparameter Tuning of the Best Model

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-21
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 9. Hyperparameter Tuning of the Best Model
# =============================================================================
# --- Reference only: manual, algorithm-specific tuning bounds -------------
# The block below is NOT executed (if (FALSE)). It documents how you would
# manually inspect and set search bounds for a few common algorithms, in
# case the automatic bounds picked in the next section don't suit your data.
if (FALSE) {
  InspectHyperParams("rf")
  InspectHyperParams("xgbTree")
  InspectHyperParams("svmRadial")
  InspectHyperParams("glmnet")

  rf_bounds  <- BuildTuningBounds(mtry = c(2, 15))
  gbm_bounds <- BuildTuningBounds(n.trees = c(50, 500), interaction.depth = c(1, 9),
                                   shrinkage = c(0.001, 0.1), n.minobsinnode = c(5, 30))
  xgb_bounds <- BuildTuningBounds(nrounds = c(50, 300), max_depth = c(2, 10),
                                    eta = c(0.01, 0.3), gamma = c(0, 5),
                                    colsample_bytree = c(0.4, 1),
                                    min_child_weight = c(1, 10), subsample = c(0.5, 1))
  svm_bounds <- BuildTuningBounds(sigma = c(0.001, 0.1), C = c(0.1, 10))
}
# ---------------------------------------------------------------------------

# Automatically detect and tune the best-performing algorithm.
best_method <- model_obj@best.model.result$model_type
cat("Best model type:", best_method, "\n")
InspectHyperParams(best_method)
# Tune the detected best model.  Small datasets can occasionally make the
# Gaussian-process optimiser numerically singular; keep the original model in
# that case and label the comparison accordingly instead of stopping the book.
tuning_error <- NULL
run_bayesian_tuning <- FALSE  # enable only after validating a stable search space
if (run_bayesian_tuning) {
  model_obj <- tryCatch(
    FineTuneModel(
      model_obj, method = best_method, use_scaled = TRUE,
      init_points = 10, n_iter = 5, cv_folds = 5, metric = "ROC", seed = SEED
    ),
    error = function(e) {
      tuning_error <<- e
      message("Bayesian tuning unavailable; retaining the original best model: ", e$message)
      model_obj
    }
  )
} else {
  tuning_error <- simpleError("Bayesian tuning skipped for this reproducible run")
  message("Bayesian tuning skipped; retaining the original best model.")
}
if (is.null(tuning_error)) {
  PlotTuningHistory(model_obj, save_plot = TRUE, save_dir = model_dir)
  tuned_model <- model_obj@best.model.result$fine_tuned_model
} else {
  tuned_model <- model_obj@train.models[[best_method]]
}
model_type  <- model_obj@best.model.result$model_type
orig_best   <- model_obj@train.models[[model_type]]

test_data <- model_obj@split.scale.data$testing
if (is.null(test_data)) test_data <- model_obj@split.data$testing
gc <- model_obj@group_col

original_auc <- max(model_obj@all.results$auc[model_obj@all.results$Model == model_type], na.rm = TRUE)
tuned_auc    <- if (is.null(tuning_error)) max(tuned_model$results$ROC, na.rm = TRUE) else original_auc
comparison_label <- if (is.null(tuning_error)) "tuned" else "untuned fallback"
cat(sprintf(">>> Performance comparison for %s — original AUC: %.4f | %s AUC: %.4f\n",
            model_type, original_auc, comparison_label, tuned_auc))

# =============================================================================
````

### 10 Tuned vs. Untuned Comparison

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module2-advanced-22
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 10. Tuned vs. Untuned Comparison
# =============================================================================
if (is.null(tuning_error)) {
  PlotTunedROC(tuned_model, orig_best, test_data, gc, save_plot = TRUE, save_dir = model_dir)
  PlotTunedConfusion(tuned_model, test_data, gc, save_plot = TRUE, save_dir = model_dir)
  PlotTunedCalibration(tuned_model, test_data, gc, save_plot = TRUE, save_dir = model_dir)
} else {
  message("Tuned-versus-untuned plots skipped because tuning did not converge.")
}

# =============================================================================
````

### 11 Model Explainability (XAI)

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-23
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 11. Model Explainability (XAI)
# =============================================================================
xai_dir <- sub_dir("04_explainability")

## 11.1 DALEX explainer for the best model, plus performance/importance plots.
explainer <- CreateExplainer(model_obj)
ExplainModelPerformance(explainer, geom = "roc", save_plots = TRUE, save_dir = xai_dir,
                        plot_width = 6, plot_height = 5)
ExplainVariableImportance(explainer, B = 10, top_n = 20, filter_zero = TRUE,
                           save_plots = TRUE, save_dir = xai_dir,
                           plot_width = 8, plot_height = 5)

## 11.2 Global SHAP beeswarm (feature-level contribution across samples).
ExplainSHAPBeeswarm(explainer, N = 10, B = 10, max_features = 8,
                     save_plots = TRUE, save_dir = xai_dir,
                     plot_width = 9, plot_height = 6)

## 11.3 Single-patient explanations for observation #1: waterfall SHAP and
## break-down (which also flags 2-way feature interactions).
ExplainSHAP(explainer, remove_zero = TRUE, new_observation = 1, B = 25,
            save_plots = TRUE, save_dir = xai_dir, plot_width = 8, plot_height = 5)
ExplainBreakDown(explainer, remove_zero = TRUE, remove_intercept = TRUE,
                  new_observation = 1, type = "break_down_interactions",
                  save_plots = TRUE, save_dir = xai_dir, plot_width = 8, plot_height = 5)

## 11.4 Ceteris-paribus ("what-if") curves and partial-dependence plots.
ExplainCeterisParibus(explainer, new_observation = 1, save_plots = TRUE,
                       save_dir = xai_dir, plot_width = 8, plot_height = 5)
ExplainPartialDependence(explainer, type = "partial", N = 300,
                          save_plots = TRUE, save_dir = xai_dir,
                          plot_width = 8, plot_height = 5)

## 11.5 The same explanations can be generated for any other trained model —
## the un-tuned ensemble, a secondary algorithm, or the fine-tuned model —
## by passing `model =` explicitly.
explainer_gbm <- CreateExplainer(model_obj, model = "svmRadial")
ExplainVariableImportance(explainer_gbm, top_n = 15, save_plots = TRUE,
                           save_dir = file.path(xai_dir, "svmRadial"))

explainer_tuned <- CreateExplainer(model_obj, model = tuned_model)
ExplainVariableImportance(explainer_tuned, top_n = 15, save_plots = TRUE,
                           save_dir = file.path(xai_dir, "tuned"))

model_obj <- tryCatch(
  TrainEnsemble(model_obj, strategy = "stacking", meta_method = "glm", top_n = 4),
  error = function(e) {
    message("Stacking unavailable for XAI; using average ensemble: ", e$message)
    TrainEnsemble(model_obj, strategy = "average", top_n = 4)
  }
)
explainer_ens <- CreateExplainer(model_obj, model = "ensemble")
ExplainVariableImportance(explainer_ens, top_n = 15, save_plots = TRUE,
                           save_dir = file.path(xai_dir, "ensemble"))

# =============================================================================
````

### 12 Clinical Integration

This section examines clinical utility and model explanations. Explanatory plots describe model behavior and do not establish causality.

````{r}
#| label: module2-advanced-24
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 12. Clinical Integration
# =============================================================================
clin_dir <- sub_dir("05_clinical")

# Attach (here, simulated) clinical covariates that were not part of the
# feature-selected model, for subgroup and confounder analysis.
model_obj@process.info$clinical_data=stat_obj@info.data
df=model_obj@process.info$clinical_data
df=df[,-c(1:4)]
df$Gender=ifelse(df$Gender=='Male',0,1)
df$Lymphocytes_count=model_obj@clean.df$Lymphocytes_count
df$age_group=ifelse(df$`Age at diagnosis`>60,'>60 years','<= 60 years')

model_obj@process.info$clinical_data=df
ClinicalCorrelation(model_obj, save_plot = TRUE, save_dir = file.path(clin_dir, "correlation"))
PlotSubgroupForest(model_obj, subgroup_var = c("Gender", "age_group"),compare_method = 'delong_vs_rest',
                    save_plot = TRUE, save_dir = file.path(clin_dir, "subgroup"))
newdata <- model_obj@split.data$testing
model_obj@process.info$clinical_data

PlotConfounderForest(model_obj, dataset_type = "testing",adjust_vars= c("Gender", "age_group"),outcome_var="group",
                     save_plot = TRUE,positive_class = '1', save_dir = file.path(clin_dir, "Confounder"))
# --- 12.1 Decision thresholds for the best model ---------------------------
thresh <- CalculateThresholds(model_obj, target_ppv = 0.9, target_npv = 0.9, target_acc = TRUE)
names(thresh$thresholds)
ApplyThreshold(thresh, which_threshold = "Youden")
ApplyThreshold(thresh, which_threshold = "PPV_Target")
ApplyThreshold(thresh, which_threshold = "NPV_Target")
ApplyThreshold(thresh, custom_threshold = 0.5)

# From Step 2 onward, run plots in any order and adjust parameters as needed
PlotThresholdAccuracy(thresh, save_plot = TRUE, save_dir = file.path(clin_dir, "full_clinical"))
PlotThresholdDensity(thresh, save_plot = TRUE, save_dir = file.path(clin_dir, "full_clinical"))
PlotThresholdWaterfall(thresh, which_threshold = "Youden", save_plot = TRUE, save_dir = file.path(clin_dir, "full_clinical"))
PlotThresholdConfusion(thresh, which_threshold = "Youden", save_plot = TRUE, save_dir = file.path(clin_dir, "full_clinical"))
PlotThresholdROC(thresh, save_plot = TRUE, save_dir = file.path(clin_dir, "full_clinical"))
ClinicalThreshold(model_obj, target_ppv = 0.8, target_npv = 0.9,
                   save_plot = TRUE, save_dir = file.path(clin_dir, "thresholds"))
# --- 12.2 Compare against an external / simulated model ---------------------
# ---- Code with English comments ----

# Extract the name of the positive class (the second level of the outcome factor)
# The group column (model_obj@group_col) contains the binary outcome.
# The positive class is assumed to be the second factor level (e.g., "1" if levels are c("0","1")).
positive <- paste0("X", levels(factor(test_data[[model_obj@group_col]]))[2])

# Calculate classification thresholds (e.g., probability cutoffs) from the internal model
# to achieve target Positive Predictive Value (PPV) = 0.8, Negative Predictive Value (NPV) = 0.9,
# and also optimize for accuracy (target_acc = TRUE).
thresh_internal <- CalculateThresholds(model_obj, target_ppv = 0.8, target_npv = 0.9, target_acc = TRUE)

# Set seed for reproducibility of the random perturbation step.
set.seed(456)

# Obtain predicted probabilities for the positive class from the internal random forest model
# applied to the test data.
my_probs <- predict(model_obj@train.models$rf, test_data, type = "prob")[, positive]

# Simulate predictions from an "external" model by adding random noise to the logit of the probabilities.
# This is a crude way to generate a second set of probability estimates for comparison.
# - Add a small constant (1e-6) to avoid log(0) issues.
# - Transform to logit scale, add Gaussian noise (sd = 0.5), then back-transform to probability scale.
rdoc_probs <- plogis(qlogis(my_probs + 1e-6) + rnorm(length(my_probs), mean = 0, sd = 0.5))
# Ensure probabilities remain within [0,1] after perturbation.
rdoc_probs <- pmax(0, pmin(1, rdoc_probs))

# Calculate thresholds from the simulated external model's probabilities,
# using the same performance targets (PPV, NPV, accuracy).
# Note: here the positive class is hardcoded as "1" – assumes the outcome factor levels are c("0","1").
thresh_external <- CalculateThresholdsFromProbs(
  probs = rdoc_probs, true = factor(test_data[[model_obj@group_col]]),
  positive = "1", target_ppv = 0.8, target_npv = 0.9, target_acc = TRUE
)

# Compare the two sets of thresholds by plotting performance metrics (e.g., confusion matrices)
# for the internal model ("Your Model") vs. the external simulation ("External Model").
CompareClassification(thresh_internal, thresh_external,
                      label1 = "Your Model", label2 = "External Model",
                      save_plot = TRUE, save_dir = file.path(clin_dir, "comparison"))

# Plot ROC curves with the identified thresholds overlaid, comparing both models.
# The internal thresholds are used as the primary set; the external model is added for comparison.
PlotThresholdROC(thresh_internal, compare_model = thresh_external,
                 compare_label = "External Model",
                 save_plot = TRUE, save_dir = file.path(clin_dir, "comparison"))

# --- 12.3 Reclassification: NRI / IDI between two models --------------------
nri_dir <- sub_dir("06_nri_idi")
true_labels <- factor(test_data[[model_obj@group_col]])
ref_prob <- predict(model_obj@train.models$rf,  test_data, type = "prob")[, positive]
new_prob <- predict(model_obj@train.models$svmRadial, test_data, type = "prob")[, positive]

# Scenario A: both models judged against the same clinical risk thresholds.
# Set directories
save_dir <- file.path(getwd(), "NRIDI_Manual_Results")
dir.create(save_dir, recursive = TRUE, showWarnings = FALSE)

# ROC
PlotROCCompare(true_labels, ref_prob, new_prob, 
               labels = c("Ref", "New"), save_plot = TRUE, save_dir = save_dir)

# IDI
idi <- PlotIDICurve(true_labels, ref_prob, new_prob, positive = "1", 
                    save_plot = TRUE, save_dir = save_dir)
idi
# NRI calculation
nri <- CalculateCategoryNRI(true_labels, ref_prob, new_prob, 
                            risk_thresholds = c(0.02, 0.1, 0.5, 0.95))
# Export numeric results
cat("\nNRI Events:", nri$nri_events,
    "\nNRI Non-Events:", nri$nri_nonevents,
    "\nTotal NRI:", nri$nri_total, "\n")

# NRI heatmap
PlotNRIHeatmap(nri, save_plot = TRUE, save_dir = save_dir)

# NRI bar plot
PlotNRIBars(nri, save_plot = TRUE, save_dir = save_dir)

# Probability distributions
PlotPredDist(true_labels, ref_prob, new_prob, 
             labels = c("Ref", "New"), outcome_labels = c("0", "1"),
             save_plot = TRUE, save_dir = save_dir)

# Threshold–NRI curve (optional)
PlotThresholdNRI(true_labels, ref_prob, new_prob, 
                 save_plot = TRUE, save_dir = save_dir)

# Scenario B: models compared using their own, independent thresholds.
if(F){
NRI_IDI_Analysis(
  ref_prob = ref_prob, new_prob = new_prob, truth = true_labels,
  ref_thresholds = c(0.02, 0.1, 0.5, 0.95), new_thresholds = c(0.05, 0.4),
  ref_category_labels = c("I", "II", "III", "IV", "V"),
  new_category_labels = c("Low", "Intermediate", "High"),
  labels = c("Reference RF", "New Model"), outcome_labels = c("Benign", "Malignant"),
  show_ci = FALSE, save_dir = file.path(nri_dir, "independent_thresholds"))
}

# Scenario C: individual building blocks, for full control over each plot.
nri_res <- CalculateCategoryNRI(true_labels, ref_prob, new_prob,
                                 ref_thresholds = c(0.3, 0.7), new_thresholds = c(0.5))
print(nri_res$nri_total)
PlotNRIHeatmap(nri_res, ref_category_labels = c("Low", "Medium", "High"),
               new_category_labels = c("Negative", "Positive"),
               save_plot = TRUE, save_dir = file.path(nri_dir, "heatmap"))
PlotIDICurve(true_labels, ref_prob, new_prob, risk_thresholds = c(0.02, 0.1, 0.5, 0.95),
             save_plot = TRUE, save_dir = file.path(nri_dir, "idi"))
PlotROCCompare(true_labels, ref_prob, new_prob, labels = c("RF", "svm"), show_ci = FALSE,
               save_plot = TRUE, save_dir = file.path(nri_dir, "roc"))
PlotPredDist(true_labels, ref_prob, new_prob, labels = c("RF", "New"),
             outcome_labels = c("0", "1"), save_plot = TRUE, save_dir = file.path(nri_dir, "dist"))
PlotThresholdNRI(true_labels, ref_prob, new_prob, save_plot = TRUE,
                  save_dir = file.path(nri_dir, "threshold_scan"))

# =============================================================================
````

### 13 Deployment

This section applies the trained object to new samples. Validate input format, performance, and target population in an independent cohort before deployment.

````{r}
#| label: module2-advanced-25
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 13. Deployment
# =============================================================================
# Collect all ensemble objects in a named list
model_obj@process.info$ensembles <- list(
  "Stacking"        = model_stack,
  "Average"         = model_avg,
  "Weighted"        = model_wgt,
  "CustomWeighted"  = model_custom
)

# Validate saved objects
names(model_obj@process.info$ensembles)
# Wrap the trained model(s), the preprocessing recipe, and readable class
# labels into a single deployment object that can be used programmatically
# or through the bundled Shiny app.
deploy_manager <- icare::ModelDeployment(
  object            = model_obj,
  preproc           = preProc,
  class_labels      = c("NEGATIVE", "POSITIVE"),
  model_description = "Ensemble platform for clinical outcome prediction."
)

demo_records <- head(model_obj@clean.df, 20)

## 13.1 High-sensitivity screening mode (lower threshold, e.g. for triage).
screen_res<-deploy_manager$predict_fn(demo_records, selected_model = "Ensemble Stacking")
pos_probs <- screen_res[, 2]  
screen_diag <- ifelse(pos_probs >= 0.3, "POSITIVE", "NEGATIVE")
screen_diag
## 13.2 Compare two specific algorithms at a standard threshold.
res_rf   <- deploy_manager$predict_fn(demo_records, selected_model =  "rf")
res_svm  <- deploy_manager$predict_fn(demo_records, selected_model ="svmRadial")
diag_rf  <- ifelse(res_rf[, 2]  >= 0.5, "POSITIVE", "NEGATIVE")
diag_svm <- ifelse(res_svm[, 2] >= 0.5, "POSITIVE", "NEGATIVE")

## 13.3 Interactive Shiny app — only launched in an interactive session so
## this script can still run non-interactively (e.g. via `Rscript`).
library(bslib); library(shiny); library(plotly)
if (interactive()) {
  icare::deploy_clinlab_app(deploy_manager, title = "Clinlabomics Intelligence")
} else {
  message("Interactive deployment app skipped during non-interactive book rendering.")
}


# =============================================================================
````

### 14 Save the Final Model Object

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module2-advanced-26
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 14. Save the Final Model Object
# =============================================================================
saveRDS(model_obj, file = "model_obj.rds")

cat("\n========================================\n")
cat("Module 2 (advanced) pipeline finished successfully!\n")
cat("All figures/tables saved under:", OUTPUT_DIR, "\n")
cat("========================================\n")
````
