# Module 3: Unsupervised Subtyping — Advanced

The advanced `Subtyping` workflow normalizes independent development and validation sets, compares K-means, latent profile analysis (LPA), and NMF, and evaluates t-SNE/UMAP structure, internal cluster quality, cross-method agreement, marker profiles, silhouettes, alluvial relationships, and deployment to new samples.

## Preventing Information Leakage

Split the data first, fit min–max parameters on the development set, and apply them unchanged to validation data. The workflow also checks numeric types and removes zero-variance features. Embeddings are exploratory: choose and validate a subtype solution using stability, agreement, marker reproducibility, and clinical usefulness rather than visual separation alone.

## Step-by-step Workflow

<!-- split-into-step-chunks -->

The workflow is divided into short stages. Each note explains the purpose, key checks, and interpretation limits. Generated plots and tables appear directly below their code chunk.

### Environment and Global Setup

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module3-advanced-01
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# =============================================================================
# icare Package Vignette — Module 3: Subtyping / Unsupervised Clustering
# ADVANCED / PUBLICATION EDITION
# =============================================================================
# Audience : Advanced users who want to compare multiple clustering methods,
#            validate them, and produce publication-style figures.
# Input    : "stat_obj.rds", produced by Module 1 (data cleaning).
#
# Pipeline covered in this script:
````

### 1 Convert the cleaned Stat object into a Subtyping object

This section creates the icare analysis object that stores data, metadata, and workflow state for all later steps.

````{r}
#| label: module3-advanced-02
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   1) Convert the cleaned Stat object into a Subtyping object
````

### 2 Normalize data and split into train/validation sets

This section preprocesses the data. Review missingness, outlier handling, and normalization before accepting the defaults.

````{r}
#| label: module3-advanced-03
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   2) Normalize data and split into train/validation sets
````

### 3 Cluster with three complementary methods: K-means, LPA (Gaussian

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module3-advanced-04
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   3) Cluster with three complementary methods: K-means, LPA (Gaussian
#      mixture), and NMF (the primary method, since it also yields a
#      re-usable model for predicting subtypes in new data)
````

### 4 Reduce dimensions (t-SNE / UMAP) for visualization

This section creates exploratory and reporting figures. Plots appear below the code and are also saved as vector PDFs for publication.

````{r}
#| label: module3-advanced-05
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   4) Reduce dimensions (t-SNE / UMAP) for visualization
````

### 5 Evaluate and cross-compare the three clustering solutions

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module3-advanced-06
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   5) Evaluate and cross-compare the three clustering solutions
````

### 6 Find subtype-specific marker features and visualize them

This section creates exploratory and reporting figures. Plots appear below the code and are also saved as vector PDFs for publication.

````{r}
#| label: module3-advanced-07
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   6) Find subtype-specific marker features and visualize them
#      (heatmaps, silhouette plot, alluvial diagram)
````

### 7 Deploy the trained NMF model for predicting subtypes in new samples

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module3-advanced-08
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   7) Deploy the trained NMF model for predicting subtypes in new samples
#
# All figures are written into structured sub-folders of OUTPUT_DIR.
# =============================================================================
# NOTE: the previous first line here was `load("../all_model.csv")`, which
# is a bug -- load() only reads .RData/.rds binary files, not CSV, and would
# error immediately (before rm(list=ls()) even runs). Removed as apparent
# leftover debug code; if you need to load model presets from a CSV, use
# read.csv() instead, or icare's own data(allmodel)/list_presets().
rm(list = ls())
library(icare)

# ---------------------------------------------------------------------------
````

### 0 CONFIG

This section defines seeds, output paths, and dependencies. Update these paths and key parameters first when using new data.

````{r}
#| label: module3-advanced-09
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 0. CONFIG
# ---------------------------------------------------------------------------
INPUT_STAT_OBJ  <- "stat_obj.rds"
OUTPUT_DIR      <- "./Module3_Advanced_Output"
GROUP_COL       <- "group"   # the outcome/group column carried over from Module 1
SEED            <- 123

dir.create(OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)
sub_dir <- function(...) {
  d <- file.path(OUTPUT_DIR, ...)
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
  d
}

# =============================================================================
````

### 1 Load Data and Convert to a Subtyping Object

This section imports and organizes the data. Check dimensions and variable coding before continuing.

````{r}
#| label: module3-advanced-10
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 1. Load Data and Convert to a Subtyping Object
# =============================================================================
stat_obj <- readRDS(INPUT_STAT_OBJ)
InspectObject(stat_obj)

sub_obj <- ConvertObject(stat_obj, to = "Subtyping")
InspectObject(sub_obj)
features=colnames(sub_obj@clean.data)[colnames(sub_obj@clean.data)!='sex']
sub_obj@clean.data <- sub_obj@clean.data[, features, drop = FALSE]
# =============================================================================
````

### 2 Normalize and Split into Train / Validation Sets

This section preprocesses the data. Review missingness, outlier handling, and normalization before accepting the defaults.

````{r}
#| label: module3-advanced-11
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 2. Normalize and Split into Train / Validation Sets
# =============================================================================
# =============================================================================
````

### 2 Split into Train / Validation Sets, then Normalize

This section preprocesses the data. Review missingness, outlier handling, and normalization before accepting the defaults.

````{r}
#| label: module3-advanced-12
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 2. Split into Train / Validation Sets, then Normalize
# =============================================================================
# IMPORTANT: split BEFORE fitting any normalization parameters — mirrors the
# pattern used in Module 2 (createDataPartition first, then preProcess() fit
# on the training split only, never on the full dataset). Fitting min-max
# parameters on the full data (train+validation combined) and only THEN
# splitting would let validation-set values leak into the scaling applied to
# the validation set itself — the opposite of what we want.
set.seed(SEED)
split <- SplitSubtypingObject(sub_obj, p = 0.7, stratify_by = GROUP_COL)

# min-max scaling keeps all features non-negative, which NMF requires.
# Fit normalization on the TRAINING split only.
# min-max scaling keeps all features non-negative, which NMF requires.
# Fit normalization on the TRAINING split only.
# IMPORTANT: explicitly pass group_col = GROUP_COL ("SWAB", configured at the
# top of this script) so the outcome column is correctly excluded from
# scaling. Sub_normalize_process()'s default group_col is the literal string
# "group" -- if left unset here, "SWAB" would NOT be recognized as the
# group column, and (being non-numeric) would silently remain in scale.data,
# later crashing NMF::nmf() deep inside with an unrelated "'x' must be
# numeric" error instead of a clear one at this step.
sub_train <- Sub_normalize_process(split$train, normalize_method = "min_max",
                                   group_col = GROUP_COL)
cat("Normalization completed. scale.data dimensions:", dim(sub_train@scale.data), "\n")

# Defensive check: confirm every remaining column really is numeric before
# it reaches K-means/LPA/NMF, all three of which require a purely numeric
# input matrix.
non_numeric_cols <- names(sub_train@scale.data)[!vapply(sub_train@scale.data, is.numeric, logical(1))]
if (length(non_numeric_cols) > 0) {
  stop("Non-numeric column(s) in scale.data: ", paste(non_numeric_cols, collapse = ", "),
       ". Check group_col above, or drop these columns from clean.data first.")
}

# Extract the normalization parameters learned on the TRAINING split only, so
# they can be re-applied identically to the validation split without letting
# validation-set information leak into its own scaling (avoids data leakage,
# analogous to Module 2's preProcess() being fit on split.data$training only).
norm_params <- Sub_extract_norm_params(sub_train, verbose = TRUE)

# Apply the training-derived parameters to the validation split — the
# validation set never contributes to its own normalization statistics.
sub_test <- Sub_apply_norm_params(split$test, norm_params = norm_params, verbose = TRUE)

cat("Train set:", nrow(sub_train@clean.data), "samples | Validation set:",
    nrow(sub_test@clean.data), "samples\n")

# Remove zero-variance columns — otherwise K-means' internal PCA step errors out.
sub_train@scale.data <- remove_constant_columns(sub_train@scale.data)
sub_train@clean.data <- sub_train@clean.data[, colnames(sub_train@scale.data), drop = FALSE]

# Keep the validation split's feature set aligned with the (possibly reduced)
# training feature set after zero-variance removal.
common_feats <- intersect(colnames(sub_train@scale.data), colnames(sub_test@scale.data))
sub_test@scale.data <- sub_test@scale.data[, common_feats, drop = FALSE]
sub_test@clean.data <- sub_test@clean.data[, common_feats, drop = FALSE]

# =============================================================================
````

### 3 Clustering — Three Complementary Methods

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module3-advanced-13
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3. Clustering — Three Complementary Methods
# =============================================================================
## 3.1 K-means, with the number of clusters chosen automatically (k up to 8).
cat("\n=== K-means Clustering ===\n")
sub_train <- Sub_kmeans_with_optimal_k(
  sub_train, use_scaled_data = TRUE, k.max = 8, palette_name = "Zissou1",
  save_plots = TRUE, save_dir = sub_dir("01_kmeans"), seed = SEED
)
print(table(sub_train@info.data$cluster_kmeans))

## 3.2 LPA (Latent Profile Analysis / Gaussian mixture) — a model-based
## alternative to K-means that allows clusters of different shapes/sizes.
cat("\n=== LPA Clustering ===\n")
library(mclust)
sub_train <- Sub_lpa_with_optimal_k(
  sub_train, use_scaled_data = TRUE, max_clusters = 3, verbose = TRUE,
  color_palette = "Darjeeling1", save_plots = TRUE, save_dir = sub_dir("02_lpa"), seed = SEED
)
cat("LPA optimal K =", sub_train@Optimal.cluster, "\n")
print(table(sub_train@info.data$cluster_lpa))

## 3.3 NMF (Non-negative Matrix Factorization) — the primary method here,
## because it produces a re-usable model that can assign new/validation
## samples to the same subtypes (steps 3.3d and 4 below).
cat("\n=== NMF Clustering ===\n")
library(NMF)
nmf_dir <- sub_dir("03_nmf")

````

### 3.3a Estimate the best rank (= number of subtypes) across a candidate range.

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module3-advanced-14
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3.3a Estimate the best rank (= number of subtypes) across a candidate range.
sub_train <- Sub_nmf_estimate(sub_train, rank_range = 2:6, nrun = 5,
                              method = "brunet", save_dir = nmf_dir, seed = SEED)
````

### 3.3b Pick the best rank from the estimation above.

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module3-advanced-15
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3.3b Pick the best rank from the estimation above.
sub_train <- Sub_nmf_best_rank(sub_train, nrun = 5, method = "brunet",
                               palette_name = "Zissou1", save_dir = nmf_dir)
cat("NMF optimal rank =", sub_train@Optimal.cluster, "\n")
````

### 3.3c Assign each training sample to a subtype.

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module3-advanced-16
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3.3c Assign each training sample to a subtype.
sub_train <- Sub_nmf_assign_subtypes(sub_train)
print(table(sub_train@info.data$cluster_nmf))

````

### 3.3d Refit with more NMF runs and save a re-usable model object, so new

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module3-advanced-17
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3.3d Refit with more NMF runs and save a re-usable model object, so new
# samples (or the validation split) can be classified consistently later.
model_dir <- sub_dir("03_nmf/model")
sub_train <- Sub_nmf_train_model(
  sub_train, best_k = sub_train@Optimal.cluster, nrun = 10, method = "brunet",
  model_name = "subtyping_nmf_model", save_dir = model_dir
)

# =============================================================================
````

### 4 Predict Subtypes on the Validation Split

This section creates training and validation sets. Restrict feature selection and parameter estimation to the training set to prevent leakage.

````{r}
#| label: module3-advanced-18
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 4. Predict Subtypes on the Validation Split
# =============================================================================
cat("\n=== Predict validation set using the trained NMF model ===\n")
sub_test <- Sub_predict_subtypes(sub_test, train_object = sub_train, method = "nmf", verbose = TRUE)
cat("Validation NMF subtype distribution:\n")
print(table(sub_test@info.data$cluster_nmf))

# =============================================================================
````

### 5 Dimensionality Reduction for Visualization (t-SNE & UMAP)

This section creates exploratory and reporting figures. Plots appear below the code and are also saved as vector PDFs for publication.

````{r}
#| label: module3-advanced-19
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 5. Dimensionality Reduction for Visualization (t-SNE & UMAP)
# =============================================================================
cat("\n=== Dimensionality reduction (t-SNE & UMAP) ===\n")
dr_dir <- sub_dir("04_dim_reduction")

sub_train <- Sub_tsne_analyse(sub_train, use_scaled_data = TRUE)
sub_train <- Sub_umap_analyse(sub_train, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")

# t-SNE colored by LPA subtype.
PlotDimReduction(sub_train, reduction = "tsne", color_by = "cluster_lpa",
                 palette_name = "Darjeeling1", save_plot = TRUE,
                 save_dir = file.path(dr_dir, "tsne"), point_size = 1.8) +
  ggplot2::labs(title = "t-SNE colored by LPA subtype")

# t-SNE colored by the clinical outcome (as a factor for discrete coloring).
sub_train@info.data[[GROUP_COL]] <- factor(sub_train@info.data[[GROUP_COL]])
PlotDimReduction(sub_train, reduction = "tsne", color_by = GROUP_COL,
                 palette_name = "Royal1", save_plot = TRUE,
                 save_dir = file.path(dr_dir, "tsne"), point_size = 1.8) +
  ggplot2::labs(title = "t-SNE colored by clinical outcome")


# =============================================================================
````

### 6 Clustering Quality Evaluation

This section evaluates quality and robustness. Interpret plots, metrics, and sample distributions together.

````{r}
#| label: module3-advanced-20
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 6. Clustering Quality Evaluation
# =============================================================================
cat("\n=== Clustering quality evaluation ===\n")
# Use a separate temporary copy per method so none of them overwrite each other.
km_eval  <- sub_train; km_eval@clustered.data$group  <- sub_train@info.data$cluster_kmeans
lpa_eval <- sub_train; lpa_eval@clustered.data$group <- sub_train@info.data$cluster_lpa
nmf_eval <- sub_train; nmf_eval@clustered.data$group <- sub_train@info.data$cluster_nmf

eval_km  <- Sub_evaluation_results(km_eval,  seed = SEED)
eval_lpa <- Sub_evaluation_results(lpa_eval, seed = SEED)
eval_nmf <- Sub_evaluation_results(nmf_eval, seed = SEED)

eval_summary <- data.frame(
  Method = c("K-means", "LPA", "NMF"),
  rbind(eval_km@evaluation_results, eval_lpa@evaluation_results, eval_nmf@evaluation_results)
)
print(eval_summary)
# =============================================================================
````

### 7 Consistency Between the Three Clustering Methods

This section evaluates quality and robustness. Interpret plots, metrics, and sample distributions together.

````{r}
#| label: module3-advanced-21
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 7. Consistency Between the Three Clustering Methods
# =============================================================================
cat("\n=== Consistency between methods ===\n")
sub_compare <- sub_train
# Prefix numeric cluster labels with "S" so they are treated as categories,
# not as an ordered numeric scale, in the comparison plot.
sub_compare@info.data$cluster_lpa    <- paste0("S", sub_compare@info.data$cluster_lpa)
sub_compare@info.data$cluster_kmeans <- paste0("S", sub_compare@info.data$cluster_kmeans)

ari_mat <- compare_clusterings(
  sub_compare, methods = c("cluster_kmeans", "cluster_lpa", "cluster_nmf"), output = "matrix"
)
cat("Adjusted Rand Index matrix:\n"); print(round(ari_mat, 3))

plot_clustering_comparison(
  sub_compare, methods = c("cluster_kmeans", "cluster_lpa", "cluster_nmf"),
  save_dir = sub_dir("05_method_comparison"), width = 5, height = 4.5, base_size = 12
)

# =============================================================================
````

### 8 Subtype-Specific Marker Features & Enhanced Visualization

This section creates exploratory and reporting figures. Plots appear below the code and are also saved as vector PDFs for publication.

````{r}
#| label: module3-advanced-22
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 8. Subtype-Specific Marker Features & Enhanced Visualization
# =============================================================================
cat("\n=== Subtype-specific differential features ===\n")
deg_dir <- sub_dir("06_differential")

## 8.1 Multi-class differential analysis (one-vs-rest Wilcoxon), using the
## LPA subtypes as the grouping variable. Any of the three cluster labels
## could be used here instead.
deg_df <- sub_train@clean.data
deg_df$subtype <- factor(sub_train@info.data$cluster_lpa)

multi_deg <- batch_Wilcoxon_MultiClass(
  mat = deg_df, group_col = "subtype", only.pos = FALSE,
  save_data = TRUE, save_dir = deg_dir
)

## 8.2 Marker heatmaps of the top differential features per subtype.
# Basic version, auto z-limit:
tryCatch(
  PlotClusterHeatmap(
    sub_train, p_cutoff = 2, deg_df = multi_deg, top_n = 5,
    save_path = file.path(deg_dir, "cluster_heatmap_basic.pdf")
  ),
  error = function(e) message("Sample-level heatmap skipped (raster backend): ", e$message)
)
# With a clinical annotation bar (outcome) alongside the heatmap:
tryCatch(
  PlotClusterHeatmap(
    sub_train, p_cutoff = 2, deg_df = multi_deg, group_by = "cluster_lpa",
    annotation_cols = tolower(GROUP_COL), top_n = 5, log_transform = TRUE,
    save_path = file.path(deg_dir, "cluster_heatmap_with_outcome.pdf"),
    annotation_palette = list(swab = c("0" = "#D32F2F", "1" = "#1976D2"))
  ),
  error = function(e) message("Annotated sample-level heatmap skipped (raster backend): ", e$message)
)
# Group-mean z-score heatmap (one column per subtype, not per sample) —
# useful for a compact summary figure.
PlotGroupMeanHeatmap(
  sub_train, deg_df = multi_deg, top_n = 5, custom_levels = c("1", "2", "3"),
  z_score_type = "row", heatmap_palette = c("#2166AC", "white", "#B2182B"),
  save_path = file.path(deg_dir, "group_mean_zscore.pdf")
)

## 8.3 Silhouette plot — how well-separated are the LPA subtypes?
PlotSilhouette(sub_train, group_by = "cluster_lpa", dist_method = "euclidean",
               palette_name = "Darjeeling1", save_plot = TRUE, save_dir = sub_dir("07_silhouette"))

## 8.4 Alluvial diagram comparing outcome, K-means, LPA, and NMF assignments
## for the same samples — a compact way to show whether methods agree.
PlotMultiAlluvial(
  sub_train, cols_list = c(tolower(GROUP_COL), "cluster_kmeans", "cluster_lpa", "cluster_nmf"),
  save_plot = TRUE, save_dir = sub_dir("08_alluvial")
)

# =============================================================================
````

### 9 Deployment: Predict Subtypes for New Samples

This section applies the trained object to new samples. Validate input format, performance, and target population in an independent cohort before deployment.

````{r}
#| label: module3-advanced-23
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 9. Deployment: Predict Subtypes for New Samples
# =============================================================================
demo_raw_data <- head(sub_train@clean.data, 20)
sub_manager   <- New_Sub_Manager(sub_train)

# Console-only prediction (no UI) for a quick sanity check.
quick_pred <- sub_manager$sub_predict(head(demo_raw_data), method = "nmf")
print(quick_pred@info.data$cluster_nmf)

# Optional metadata shown in the deployment app.
my_intro <- list(
  abstract = "Cohort-scale subtyping tool.",
  citation = "Add your citation here."
)
my_var_dict <- data.frame(
  Feature     = c("gender", "age", "wbc", "platelets", "crp", "ldh"),
  Description = c("Patient biological sex (1 = Male, 2 = Female)",
                  "Patient age at diagnosis",
                  "White blood cell count",
                  "Total platelet count",
                  "C-reactive protein (inflammation marker)",
                  "Lactate dehydrogenase (tumor-burden marker)"),
  Units       = c("Category", "Years", "10^9/L", "10^9/L", "mg/L", "U/L"),
  stringsAsFactors = FALSE
)

# The interactive Shiny app is only launched in an interactive session.
if (interactive()) {
  launch_sub_deploy_app(sub_manager, var_dict = my_var_dict,
                        title = "Stratification Portal", project_info = my_intro)
}

# =============================================================================
````

### 10 Save Final Objects

This section saves model objects, tables, and figures for reproducibility, review, and reuse.

````{r}
#| label: module3-advanced-24
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 10. Save Final Objects
# =============================================================================
saveRDS(sub_train, file = file.path(OUTPUT_DIR, "final_subtyping_object.rds"))

cat("\n========================================\n")
cat("Module 3 (advanced) pipeline finished successfully!\n")
cat("Optimal cluster numbers — K-means:", eval_km@evaluation_results$n_clusters,
    "| LPA:", eval_lpa@evaluation_results$n_clusters,
    "| NMF:", eval_nmf@evaluation_results$n_clusters, "\n")
cat("Best method by Silhouette score:", eval_summary$Method[which.max(eval_summary$Silhouette)], "\n")
cat("All results saved under:", OUTPUT_DIR, "\n")
cat("========================================\n")
````
