# Module 4: Survival and Prognostic Modeling — Advanced

Using the PMID37633276 example data, the advanced `PrognosiX` workflow keeps development and validation cohorts separate and covers univariable Cox and LASSO-based feature selection, survival-learner benchmarking, tuning, Kaplan–Meier risk groups, subgroup analysis, time-dependent AUC, calibration, nomograms, robustness checks, decision curves, SHAP, and deployment.

## Important Notes

The optional development check runs only when the referenced local package path exists; GitHub and installed-package users can leave it skipped. Main settings are in `OUTPUT_ROOT`, the analysis version, outcome definitions, and the data-import section. Fix the prediction horizons and clinical decision thresholds in the analysis plan before examining validation results.

## Step-by-step Workflow

<!-- split-into-step-chunks -->

The workflow is divided into short stages. Each note explains the purpose, key checks, and interpretation limits. Generated plots and tables appear directly below their code chunk.

### Environment and Global Setup

This section performs the core analysis for this stage. Review the output immediately below the code and confirm that data ranges, assumptions, and results match the study design.

````{r}
#| label: module4-advanced-01
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# =============================================================================
# icare Package Vignette — Module 4: Survival / Prognosis Modeling (PrognosiX)
# ADVANCED / PUBLICATION EDITION — STRICT TRAIN/VALIDATION SPLIT
# =============================================================================
# Audience : Advanced users building and validating a clinical survival model
#            with all diagnostic and explanatory tools.
# Dataset  : Custom Excel file from PMID37633276_DIA_plasmaproomic.xlsx.
#            Replace the file path with your own data.
# Pipeline (all steps performed on TRAINING set, except final validation):
````

### 1 Build Stat object with time/status in info.data, convert to PrognosiX

This section creates the icare analysis object that stores data, metadata, and workflow state for all later steps.

````{r}
#| label: module4-advanced-02
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   1) Build Stat object with time/status in info.data, convert to PrognosiX
````

### 2 Feature selection (multi-method, ONLY on training set)

This section selects candidate features. Consider effect size, stability, and biological plausibility rather than a single significance measure.

````{r}
#| label: module4-advanced-03
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   2) Feature selection (multi-method, ONLY on training set)
````

### 3 Train/Validation split, algorithm benchmarking, hyperparameter tuning

This section creates training and validation sets. Restrict feature selection and parameter estimation to the training set to prevent leakage.

````{r}
#| label: module4-advanced-04
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   3) Train/Validation split, algorithm benchmarking, hyperparameter tuning
````

### 4 Model evaluation: KM curves, subgroup forest, time‑dependent AUC,

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-advanced-05
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   4) Model evaluation: KM curves, subgroup forest, time‑dependent AUC,
#      calibration, nomogram, external validation
````

### 5 Robustness checks: stability, ablation, sensitivity

This section tests sensitivity to sampling, censoring, and feature changes. Robust conclusions should remain similar under reasonable perturbations.

````{r}
#| label: module4-advanced-06
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   5) Robustness checks: stability, ablation, sensitivity
````

### 6 Decision curve analysis (DCA) and SHAP explanations (on validation set)

This section examines clinical utility and model explanations. Explanatory plots describe model behavior and do not establish causality.

````{r}
#| label: module4-advanced-07
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   6) Decision curve analysis (DCA) and SHAP explanations (on validation set)
````

### 7 Save final PrognosiX object

This section saves model objects, tables, and figures for reproducibility, review, and reuse.

````{r}
#| label: module4-advanced-08
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   7) Save final PrognosiX object
````

### 8 Deployment: prediction functions + interactive Shiny app

This section applies the trained object to new samples. Validate input format, performance, and target population in an independent cohort before deployment.

````{r}
#| label: module4-advanced-09
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
#   8) Deployment: prediction functions + interactive Shiny app
# =============================================================================
#devtools::document("../../icare-git/")
if (dir.exists("../../icare-git/")) devtools::check("../../icare-git/")
#devtools::install("../../icare-git/")
library(icare)
# ---------------------------------------------------------------------------
````

### 0 CONFIG & Environment Setup

This section defines seeds, output paths, and dependencies. Update these paths and key parameters first when using new data.

````{r}
#| label: module4-advanced-10
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 0. CONFIG & Environment Setup
# ---------------------------------------------------------------------------
set.seed(2025)

# Define version and output root
VERSION <- "Advanced"
OUTPUT_ROOT <- "./PrognosiX_Output"
OUTPUT_DIR <- file.path(OUTPUT_ROOT, VERSION)

# Helper to create subdirectories
sub_dir <- function(...) {
  d <- file.path(OUTPUT_DIR, ...)
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
  d
}

# Create version root and standard subfolders
sub_dir()
sub_dir("data")
sub_dir("01_feature_selection")
sub_dir("02_model")
sub_dir("03_evaluation")
sub_dir("04_robustness")
sub_dir("05_xai")

# Load required packages
library(icare)
required_pkgs <- c("mlr3", "mlr3proba", "mlr3viz", "mlr3learners", 
                   "survival", "survminer", "ggplot2", "caret", "dplyr",
                   "tidyr", "rms", "risksetROC", "survex", "dcurves",
                   "readxl", "patchwork", "ggrepel", "maxstat")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs)) install.packages(missing_pkgs)
invisible(lapply(required_pkgs, library, character.only = TRUE))

# ---------------------------------------------------------------------------
````

### 1 Data Preparation (custom Excel data)

This section imports and organizes the data. Check dimensions and variable coding before continuing.

````{r}
#| label: module4-advanced-11
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 1. Data Preparation (custom Excel data)
# ---------------------------------------------------------------------------
raw_dis <- as.data.frame(read_excel("data/PMID37633276_DIA_plasmaproomic.xlsx", sheet = 3))
raw_dis <- raw_dis[!grepl(x = raw_dis$`Case ID`, pattern = "^KC"), ]

# Extract numeric features
mat <- raw_dis[, c("White blood cell count", 
                   "Lymphocytes%", "Monocytes%", "Neutrophilic granulocyte%", "Eosinophil%", 
                   "Basophil%", "Lymphocytes count", "Monocytes count", "Neutrophilic granulocyte count", 
                   "Eosinophil count", "Basophil count", "Red blood cell count", 
                   "Hemoglobin", "Hematocrit", "Mean red blood cell volume", "Mean cell hemoglobin", 
                   "Mean corpuscular hemoglobin concentration", "Red blood cell distribution width CV", 
                   "Red blood cell distribution width SD", "Platelet count", "Mean platelet volume", 
                   "PCT", "Platelet volume distribution width")]

index <- grepl(x = raw_dis$`Case ID`, pattern = "^HC")

inf <- raw_dis[, c("Case ID", "Sample type", "Plasma proteome ID", "Tissue proteome ID", 
                   "Gender", "Age at diagnosis", "Tobacco smoking history", "Drinking history", 
                   "Vital status", "Days to last followup", "Days until death", 
                   "Combined days to last followup or death", "Progressive-free survival_Status", 
                   "Progressive-free survival_Months", "Tumor location", "Tumor size (cm)", 
                   "T-category", "Lymph node involvement", "Distant metastasis", 
                   "TNM Stage", "Grade", "Morphology")]

inf$group <- ifelse(index, '0', '1')
mat$group <- inf$group
mat <- as.matrix(mat)
storage.mode(mat) <- "numeric"

rownames(mat) <- inf$`Case ID`
rownames(inf) <- inf$`Case ID`
mat <- as.data.frame(mat)
mat$age <- inf$`Age at diagnosis`
mat$sex <- inf$Gender

inf$time <- inf$`Combined days to last followup or death`
inf$event <- inf$`Vital status`
inf$event <- ifelse(inf$event == "Dead", 0, 1)

inf <- inf[inf$group == 1, ]
mat <- mat[rownames(inf), ]
mat$time <- inf$time
mat$status <- inf$event

stat_obj <- CreateStatObject(
  clean.data = mat,
  info.data = inf,
  group_col = "group",
  na.action = "allow"
)
cat("Stat object created.\n")

prog <- Stat_to_PrognosiX(stat_obj, "time", "status",
                          na_action = "omit",
                          min_events = 10,
                          verbose = TRUE)

# ---------------------------------------------------------------------------
````

### 2 Feature Selection (ONLY on training set)

This section selects candidate features. Consider effect size, stability, and biological plausibility rather than a single significance measure.

````{r}
#| label: module4-advanced-12
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 2. Feature Selection (ONLY on training set)
# ---------------------------------------------------------------------------
task_full <- surv_extract_task(prog)
n <- task_full$nrow
train_idx <- caret::createDataPartition(task_full$data()[[task_full$target_names[2]]],
                                        p = 0.7, list = FALSE)
val_idx <- setdiff(seq_len(n), train_idx)

train_task <- task_full$clone()$filter(train_idx)
val_task <- task_full$clone()$filter(val_idx)

methods_to_run <- c("uni_cox", "lasso", "rf_imp", "ridge", "enet", "vimp")
feat_sel <- surv_feature_selection_multi(
  object = train_task,
  methods = methods_to_run,
  p_threshold = 0.1,
  top_ratio = 0.5,
  combine = "union",
  verbose = TRUE
)
selected_feats <- feat_sel$selected
cat("Selected features:", paste(selected_feats, collapse = ", "), "\n")
write.csv(feat_sel$method_table,
          file = file.path(sub_dir("01_feature_selection"), "feature_selection_methods.csv"),
          row.names = FALSE)

train_task <- train_task$select(selected_feats)
val_task <- val_task$select(selected_feats)

# Save train/validation data
train_data <- as.data.frame(train_task$data())
val_data <- as.data.frame(val_task$data())
write.csv(train_data, file = file.path(sub_dir("data"), "training_data.csv"), row.names = FALSE)
write.csv(val_data,   file = file.path(sub_dir("data"), "validation_data.csv"), row.names = FALSE)

# ---------------------------------------------------------------------------
````

### 3 Algorithm Benchmarking & Hyperparameter Tuning

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-advanced-13
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 3. Algorithm Benchmarking & Hyperparameter Tuning
# ---------------------------------------------------------------------------
learners_list <- list(
  surv_get_learner("surv.coxph",     train_task),
  surv_get_learner("surv.cv_glmnet", train_task),
  surv_get_learner("surv.ranger",    train_task)
)
bmr1 <- surv_run_algorithm_benchmark(train_task, learners_list)
dev.new()
bmr2 <- surv_benchmark_learners(train_task, learner_ids = c("surv.coxph", "surv.ranger"), tune = FALSE)
bmr2_summary <- surv_summarize_benchmark(bmr2)
write.csv(bmr2_summary, file = file.path(sub_dir("02_model"), "algorithm_benchmark.csv"), row.names = FALSE)

best_id <- "surv.ranger"
tune <- surv_train_and_tune(train_task, best_id, tuning_budget = 25)
best_lrn <- tune$learner
cv_cindex <- tune$cv_performance
cat("CV C-index after tuning:", round(cv_cindex, 4), "\n")

# ---------------------------------------------------------------------------
````

### 4 Model Evaluation

This section trains and compares candidate models. Select models using discrimination, calibration, stability, and clinical utility together.

````{r}
#| label: module4-advanced-14
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 4. Model Evaluation
# ---------------------------------------------------------------------------
best_lrn$predict_type <- "distr"
train_perf <- surv_evaluate_model(best_lrn, train_task,
                                  measures = list(msr("surv.cindex"), msr("surv.graf")))
cat("Training apparent C-index:", round(train_perf$surv.cindex, 4), "\n")

val_pred <- best_lrn$predict(val_task)
val_cindex <- val_pred$score(msr("surv.cindex"))
cat("Validation C-index:", round(val_cindex, 4), "\n")

# Save performance metrics
perf_df <- data.frame(
  Dataset = c("Training", "Validation"),
  C_index = c(train_perf$surv.cindex, val_cindex)
)
write.csv(perf_df, file = file.path(sub_dir("03_evaluation"), "performance.csv"), row.names = FALSE)

# ---- 4.1 KM Curves ----
# IMPORTANT: The following KM curves and the cutoff derived from the training set
# are APPARENT (optimistic) performance. They are useful for internal exploration
# but MUST be re‑evaluated on an independent validation set before reporting.
pred_train <- best_lrn$predict(train_task)
cut_df <- as.data.frame(train_task$data())
cut_df$risk <- pred_train$crank

# Compute optimal cutoff using multiple methods.
# For publication, consider using 'maxstat' (Hothorn & Lausen, 2003) with
# bootstrap resampling to obtain a stable, data‑driven threshold.
# The 'p_optimize' method is also provided for historical consistency.
cut_res <- tryCatch({
  surv_cutpoint(cut_df, time = "time", event = "status", variables = "risk")
}, error = function(e) {
  warning("surv_cutpoint failed: ", e$message, ". Using median split.")
  list(cutpoint = data.frame(cutpoint = median(cut_df$risk, na.rm = TRUE)))
})
opt_cutoff <- cut_res$cutpoint$cutpoint
cat("Optimal cutoff (from surv_cutpoint):", round(opt_cutoff, 4), "\n")

# Generate KM plots for different cutoff methods
# Note: 'p_optimize' and 'maxstat' use bootstrap/maximally selected statistics
# and are exploratory; their p‑values should not be used as confirmatory.
km_median    <- surv_plot_risk_km(best_lrn, train_task, "median", risk_table = TRUE)
km_tertile   <- surv_plot_risk_km(best_lrn, train_task, "tertile")
km_quartile  <- surv_plot_risk_km(best_lrn, train_task, "quartile")
km_p_optimize <- surv_plot_risk_km(best_lrn, train_task, "p_optimize",
                                   n_boot = 100, fraction = 0.7)
km_maxstat   <- surv_plot_risk_km(best_lrn, train_task, "maxstat",
                                  n_boot = 100, minprop = 0.3)

# Save training KM plots
pdf(file.path(sub_dir("03_evaluation"), "KM_train_median.pdf"), width = 8, height = 7)
print(km_median); dev.off()
pdf(file.path(sub_dir("03_evaluation"), "KM_train_tertile.pdf"), width = 8, height = 7)
print(km_tertile); dev.off()
pdf(file.path(sub_dir("03_evaluation"), "KM_train_quartile.pdf"), width = 8, height = 7)
print(km_quartile); dev.off()
pdf(file.path(sub_dir("03_evaluation"), "KM_train_p_optimize.pdf"), width = 8, height = 7)
print(km_p_optimize); dev.off()
pdf(file.path(sub_dir("03_evaluation"), "KM_train_maxstat.pdf"), width = 8, height = 7)
print(km_maxstat); dev.off()

# Apply a fixed cutoff (here we use the median from training) to the validation set
train_cutoff <- get_cf(km_median)
if (is.null(train_cutoff)) train_cutoff <- opt_cutoff
km_val_fixed <- surv_plot_risk_km(best_lrn, val_task,
                                  cutoff_method = "custom",
                                  custom_cutoffs = train_cutoff,
                                  risk_table = TRUE)
pdf(file.path(sub_dir("03_evaluation"), "KM_val_using_train_cutoff.pdf"), width = 8, height = 7)
print(km_val_fixed); dev.off()

# ---- 4.2 Subgroup Forest (FIXED: use subsetted prog) ----
# Subset prog to training samples to match train_task rows
train_prog <- prog
train_prog@clean.data <- prog@clean.data[train_idx, , drop = FALSE]
train_prog@info.data <- prog@info.data[train_idx, , drop = FALSE]
train_prog@survival.data <- prog@survival.data[train_idx, , drop = FALSE]

clinical_cols <- colnames(train_prog@info.data)
candidate_vars <- c("Gender", "Tobacco.smoking.history", "Drinking.history", "TNM.Stage")
subgroup_vars <- intersect(candidate_vars, clinical_cols)

forest <- surv_plot_subgroup_forest(
  learner = best_lrn,
  object = train_task,
  subgroup_vars = subgroup_vars,
  prog = train_prog,
  save_plot = TRUE,
  save_dir = sub_dir("03_evaluation"),
  file_name = "subgroup_forest.pdf"
)
print(forest)

# ---- 4.3 Time-dependent AUC ----
# Note: These are also apparent performance when computed on the training set.
auc_train <- surv_plot_time_dependent_auc(best_lrn, train_task)
auc_val   <- surv_plot_time_dependent_auc(best_lrn, val_task)
auc_cmp   <- surv_plot_comparison_auc(best_lrn, train_task, val_task)
ggsave(file.path(sub_dir("03_evaluation"), "AUC_comparison.pdf"), auc_cmp, width = 8, height = 5)

# ---- 4.4 Calibration ----
cal_train <- surv_plot_calibration(best_lrn, train_task, time_point = 200)
cal_cmp   <- surv_plot_comparison_calibration(best_lrn, train_task, val_task, time_point = 200)
ggsave(file.path(sub_dir("03_evaluation"), "Calibration_train.pdf"), cal_train, width = 8, height = 6)
ggsave(file.path(sub_dir("03_evaluation"), "Calibration_comparison.pdf"), cal_cmp, width = 8, height = 6)

# ---- 4.5 External Validation (using validation set) ----
val_data_ext <- as.data.frame(task_full$data())[val_idx, selected_feats, drop = FALSE]
val_data_ext$time <- task_full$data()[[task_full$target_names[1]]][val_idx]
val_data_ext$status <- task_full$data()[[task_full$target_names[2]]][val_idx]
val_res <- surv_predict_on_validation(best_lrn, val_data_ext, train_task)
val_cindex_alt <- val_res$prediction$score(msr("surv.cindex"))
cat("Validation C-index (alt):", round(val_cindex_alt, 4), "\n")

# ---- 4.6 Nomogram ----
# The revized surv_generate_nomogram now includes a proportional hazards test
# (cox.zph) which will be printed to the console. If any variable violates
# the PH assumption, a warning is shown.
nom <- surv_generate_nomogram(train_task,
                              selected_features = head(selected_feats, 5),
                              time_points = c(90, 180, 365),
                              time_unit = "days")
# Nomogram is drawn; optionally save via pdf().

# ---------------------------------------------------------------------------
````

### 5 Robustness Analysis

This section tests sensitivity to sampling, censoring, and feature changes. Robust conclusions should remain similar under reasonable perturbations.

````{r}
#| label: module4-advanced-15
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 5. Robustness Analysis
# ---------------------------------------------------------------------------
stab <- surv_analyze_feature_stability(train_task, "time", "status", n_repeat = 20, alpha = 1)
cat("Stability index:", round(stab$stability_index, 4), "\n")
ggsave(file.path(sub_dir("04_robustness"), "stability_plot.pdf"), stab$plot, width = 8, height = 6)
write.csv(stab$frequencies, file.path(sub_dir("04_robustness"), "stability_frequencies.csv"), row.names = FALSE)

abl <- surv_analyze_feature_ablation(train_task, best_id, selected_feats)
ggsave(file.path(sub_dir("04_robustness"), "ablation_plot.pdf"), abl$plot, width = 8, height = 6)
write.csv(abl$results, file.path(sub_dir("04_robustness"), "ablation_results.csv"), row.names = FALSE)

sens_sample <- surv_analyze_model_sensitivity(train_task, best_id, analysis_type = "sample_size")
sens_censor <- surv_analyze_model_sensitivity(train_task, best_id, analysis_type = "censoring")
ggsave(file.path(sub_dir("04_robustness"), "sensitivity_sample.pdf"), sens_sample$plot, width = 8, height = 5)
ggsave(file.path(sub_dir("04_robustness"), "sensitivity_censor.pdf"), sens_censor$plot, width = 8, height = 5)
write.csv(sens_sample$results, file.path(sub_dir("04_robustness"), "sensitivity_sample.csv"), row.names = FALSE)
write.csv(sens_censor$results, file.path(sub_dir("04_robustness"), "sensitivity_censor.csv"), row.names = FALSE)

# ---------------------------------------------------------------------------
````

### 6 DCA & SHAP

This section examines clinical utility and model explanations. Explanatory plots describe model behavior and do not establish causality.

````{r}
#| label: module4-advanced-16
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 6. DCA & SHAP
# ---------------------------------------------------------------------------
val_time <- val_task$data()[[val_task$target_names[1]]]
val_status <- val_task$data()[[val_task$target_names[2]]]
event_times <- val_time[val_status == 1]
if (length(event_times) == 0) stop("No events in validation set.")
eval_time <- round(quantile(event_times, 0.5, na.rm = TRUE), 0)
if (eval_time < min(event_times)) eval_time <- min(event_times)
cat("Using eval_time =", eval_time, "for DCA\n")

dca_res <- plot_dca_survival(
  learners = list("Tuned" = best_lrn),
  object = val_task,
  eval_time = eval_time,
  clin_range = c(0.05, 0.5),
  print_stats = TRUE
)

ggsave(file.path(sub_dir("05_xai"), "dca_plot.pdf"), dca_res$plot, width = 7, height = 5)
write.csv(dca_res$table, file.path(sub_dir("05_xai"), "dca_table.csv"), row.names = FALSE)

shap_global <- surv_explain_shap(
  learner = best_lrn,
  task = val_task,
  type = "global",
  n_explain = 30,
  n_background = 10,
  n_timepoints = 10,
  n_top_features = 6,
  verbose = TRUE
)
ggsave(file.path(sub_dir("05_xai"), "shap_global_bar.pdf"), shap_global$plots$bar_plot, width = 8, height = 6)
if (!is.null(shap_global$plots$line_plot)) {
  ggsave(file.path(sub_dir("05_xai"), "shap_global_line.pdf"), shap_global$plots$line_plot, width = 8, height = 6)
}
write.csv(shap_global$shap_long, file.path(sub_dir("05_xai"), "shap_values_global.csv"), row.names = FALSE)

beeswarm_plot <- surv_plot_shap_beeswarm(shap_global, top_n = 6, method = "beeswarm",
                                         title = "Global SurvSHAP (validation set)")
ggsave(file.path(sub_dir("05_xai"), "shap_beeswarm.pdf"), beeswarm_plot, width = 9, height = 6)

# ---------------------------------------------------------------------------
````

### 7 Save Final PrognosiX Object

This section saves model objects, tables, and figures for reproducibility, review, and reuse.

````{r}
#| label: module4-advanced-17
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 7. Save Final PrognosiX Object
# ---------------------------------------------------------------------------
prog@best.model <- list(
  learner_id = best_id,
  learner = best_lrn,
  best_params = tune$best_params,
  cv_cindex = cv_cindex,
  features = selected_feats,
  cutoff = opt_cutoff,
  decision_type = "binary",
  train_cols = train_task$feature_names
)
prog@subgroup.risk <- list(
  benchmark_table = bmr1$table,
  stability = stab,
  ablation = abl,
  validation = list(cindex = val_cindex, n = val_task$nrow)
)
prog@split.data <- list(train_idx = train_idx, val_idx = val_idx)
saveRDS(prog, file = file.path(sub_dir("data"), "prog_obj_final.rds"))

# ---------------------------------------------------------------------------
````

### 8 Deployment

This section applies the trained object to new samples. Validate input format, performance, and target population in an independent cohort before deployment.

````{r}
#| label: module4-advanced-18
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 8. Deployment
# ---------------------------------------------------------------------------
# ---------------------------------------------------------------------------
````

### 8 Deployment: Prediction for New Samples

This section applies the trained object to new samples. Validate input format, performance, and target population in an independent cohort before deployment.

````{r}
#| label: module4-advanced-19
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 8. Deployment: Prediction for New Samples
# ---------------------------------------------------------------------------
# Instead of the veteran example, use actual rows from your training data
# (or real new data) that contain all selected features.
# For demonstration, we take the first 2 rows of the original feature matrix.
new_patients <- data.frame(
  Mean_corpuscular_hemoglobin_concentration = c(33, 34),
  Basophil_count = c(0.02, 0.03),
  Monocytes = c(0.6, 0.7),
  Neutrophilic_granulocyte = c(4.2, 5.1),
  Red_blood_cell_distribution_width_CV = c(12.3, 11.8),
  Mean_red_blood_cell_volume = c(89, 92),
  Lymphocytes_count = c(1.8, 2.1),
  Platelet_volume_distribution_width = c(12.0, 13.1),
  Lymphocytes = c(1.8, 2.1),   # note: this may be same as Lymphocytes_count? check your data
  Monocytes_count = c(0.6, 0.7),
  Neutrophilic_granulocyte_count = c(4.2, 5.1),
  Mean_platelet_volume = c(10.5, 11.0),
  row.names = c("patient_1", "patient_2")
)

risk_scores <- predict_prognosix(prog, new_patients, impute = TRUE)
pred_groups_med <- predict_risk_groups(prog, new_patients, cutoff_method = "median")
pred_groups_cus <- predict_risk_groups(prog, new_patients,
                                       cutoff_method = "custom",
                                       custom_cutoffs = opt_cutoff)
cat("Risk scores:\n"); print(risk_scores)
cat("Median groups:\n"); print(pred_groups_med)
cat("Custom groups:\n"); print(pred_groups_cus)

manager <- New_Prog_Manager(prog)

if (interactive()) {
  use_app_theme_grey()
  set_prog_app_text(
    title = "Survival Risk Predictor",
    citation_text = "Data: PMID37633276_DIA_plasmaproomic cohort."
  )
  set_prog_app_theme(
    primary_color = "#2c7fb8", background_color = "#f0f4f8", sidebar_color = "#e9ecef",
    box_background = "#ffffff", label_color = "#2c7fb8",
    run_button_gradient_start = "#2c7fb8", run_button_gradient_end = "#1d4e6e",
    risk_high_color = "#d9534f", risk_medium_color = "#f0ad4e", risk_low_color = "#5cb85c",
    table_header_color = "#2c7fb8", font_family = "Arial, sans-serif", font_size_base = 14
  )
  launch_prog_deploy_app(
    manager,
    var_dict = data.frame(
      Feature = selected_feats,
      Description = paste("Predictor", seq_along(selected_feats)),
      Units = rep("--", length(selected_feats))
    ),
    project_info = list(
      abstract = paste("Prognostic model trained on", train_task$nrow, "patients."),
      citation = "PMID37633276_DIA_plasmaproomic cohort."
    )
  )
}

# ---------------------------------------------------------------------------
````

### 9 Final Report

This section saves model objects, tables, and figures for reproducibility, review, and reuse.

````{r}
#| label: module4-advanced-20
#| cache: false
#| echo: true
#| message: false
#| warning: false
#| fig-show: hold
#| out-width: '100%'
# 9. Final Report
# ---------------------------------------------------------------------------
cat("\n========================================\n")
cat("Module 4 (advanced, leakage‑free) finished successfully!\n")
cat("Best model:", best_id,
    "| CV C-index (training):", round(cv_cindex, 4),
    "| Validation C-index:", round(val_cindex, 4), "\n")
cat("Optimal cutoff:", round(opt_cutoff, 4), "\n")
cat("All results saved under:", OUTPUT_DIR, "\n")
cat("========================================\n")
````

## Key Figure Preview

### SCI Composite: Survival Model Performance and Explanation

Panels show: A, training Kaplan–Meier; B, validation Kaplan–Meier; C, time-dependent AUC; D, calibration; E, subgroup effects; F, survival SHAP. Distinguish training from validation and discuss risk stratification, discrimination, calibration, heterogeneity, and explanation.

<object data="image/composite_figures/module4_composite.pdf" type="application/pdf" width="100%" height="760"><a href="image/composite_figures/module4_composite.pdf">Download Module 4 SCI composite</a></object>

<object data="image/module4/key_figures/km_train.pdf" type="application/pdf" width="100%" height="650"><a href="image/module4/key_figures/km_train.pdf">Training-set Kaplan–Meier curve</a></object>

<object data="image/module4/key_figures/km_validation.pdf" type="application/pdf" width="100%" height="650"><a href="image/module4/key_figures/km_validation.pdf">Validation-set Kaplan–Meier curve</a></object>

<object data="image/module4/key_figures/km_advanced_train.pdf" type="application/pdf" width="100%" height="650"><a href="image/module4/key_figures/km_advanced_train.pdf">Advanced-model training Kaplan–Meier curve</a></object>
